---
title: "week06-probability"
output: html_document
date: "2024-10-07"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("~/Desktop/STAT 240/scripts/ggprob.R")
```

```{r}
data = tibble(x = 2:12, `P(X=x)` = round(c(1:6, 5:1)/36, 2)) %>% print(n = Inf)
```

Mean: $\mu = \sum_x x \cdot P(X=x)$

```{r}
data %>% mutate(xP = x*`P(X=x)`) %>% summarise(mean = sum(xP))

mean = sum(data$x * data$`P(X=x)`)
mean

mean(data$`P(X=x)`) # is not that the same as mean/expected value. The mean is just the average
```

Variance: $\sigma^2 = \sum_x (x-\mu)^2 \cdot P(X=x)$

```{r}
v = data %>% mutate(v = ((x-mean)^2)*`P(X=x)`) %>% summarise(variance = sum(v))
variance = v$variance

variance = sum( (data$x - mean)^2 * data$`P(X=x)`)
variance
```

Standard Deviation: $\sigma = \sqrt(variance)$

```{r}
standard_deviation = sqrt(variance)
standard_deviation
```





# Sec 003 Monday 10/14

Discrete R.V

mean $\mu$, $\mu = \sum_{x}x \cdot P(X= x)$

```{r}
X = tibble(x = 2:12, `P(X=x)` = round(c(1:6, 5:1)/36, 2)) %>% print(n = Inf)
X %>% mutate(xP = x*`P(X=x)`) %>% summarise(mean = sum(xP))
mean = sum(X$x * X$`P(X=x)`)
mean

mean(X$`P(X=x)`) # not the same as the expected value/mean. Since it only finds the average
```

Variance: $\sigma^2 = \sum_x (x-\mu)^2 \cdot P(X= x)$ 

```{r}
X %>% mutate(v = (x-mean)^2*`P(X=x)`) %>% summarise(variance = sum(v))

variance = sum((X$x - mean)^2 * X$`P(X=x)`)
variance
```

Standard deviation: $\sigma = \sqrt(variance)$

```{r}
deviation = sqrt(variance)
deviation
```



# Sec 006-Monday 10/14

Discrete Distribution:

Mean: $\mu$

$\mu = \sum x P(X=x)$

Variance: $\sigma^2$

$\sigma^2 = \sum (x-\mu)^2 P(X=x)$

Standard Deviation: $\sigma$

$\sigma = \sqrt(\sigma^2)$

```{r}
x = tibble(x = 2:12, `P(X=x)` = round(c(1:6, 5:1)/36, 2)) %>% print(n = Inf)
mean = sum(x$x * x$`P(X=x)`)
variance = sum((x$x-sum(x$x * x$`P(X=x)`))^2 * x$`P(X=x)`)
standard_deviation = sqrt(variance)
```

# Sec 005 - Monday 10/7

Mathematical expression

$$\frac{1}{n}\sum (x_i - \mu)^2$$

# Sec 003 - Monday 10/7

Mathematical expression

$$\frac{1}{n}\sum (x_{i} - \mu)^2$$

# Sec 006-Monday 10/7

Mathematical expression

$\frac{1}{n}\sum (x_i-\mu)$
