---
title: "Discussion 11"
output: html_document
---

# Preliminaries

-   This file should be in `STAT240/discussion/ds11` on your local computer.

-   This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

-   This assignment is to be completed in the first 10-20 mins of the discussion session. The rest of the discussion time the students are to work on the group projects with their discussion groups.

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Members:

> Absent Group Members:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
```

## Problems

The questions in this discussion assignment use the chimpanzee data set.
In answering these questions, you may assume the following in models for the data:

1. all trials are independent
2. trials made with a partner have the same chance of being pro-social, even if the identities of the actors or partners differ.

## Problem 1

**Part(a)** Using the chimpanzee data, combine all of the trials made without a partner.

Count the number of pro-social and selfish choices made.

```{r}
# Write your code here! 
```

**Part(b)**

Write a model for the number of pro-social choices.

> Write your answer here! 


**Part(c)**

Use the Agresti-Coull method to construct a 95% confidence interval for the probability that a chimpanzee makes the pro-social choice in experimental conditions when there is no partner present.

```{r}
# Write your code here! 
```


**Part(d)**

Interpret this confidence interval in the context of the experimental setting.

> Write your answer here! 

## Problem 2

Using the same data as in the previous problem,
do the following.

**Part(a)**

State hypotheses for a two-sided hypothesis test that the probability that a chimpanzee makes the pro-social choice over the long run is equal to 0.5.

> Write your answer here! 


**Part(b)**

Pick a test statistic, state the sampling distribution of this statistic if the null hypothesis is true, and compute a p-value for the test.

```{r}
# Write your code here! 
```

**Part(c)**

Interpret these results in the context of the experiment following the examples from lecture notes.

> Write your answer here! 

### Please use rest of your time to work on the project proposal in your groups. Come up with an appropriate dataset, a question to ask, and some ideas of how to proceed. 
