---
title: "ggprob"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```

Professor Bret Larget has written several shortcut functions in the file `ggprob.R`  Just like how we load a package like `tidyverse`, we can load commands from a different R file on your computer.

- Download this .Rmd to `STAT240/lectures/extras`
- Download `ggprob.R` to `STAT240/scripts`

The functions in `ggprob` can help us quickly create plots of different distributions.

```{r}
source("../../scripts/ggprob.R")
```

For example, let's plot the distribution of a Binom(100, 0.5) variable.

```{r}
gbinom(100, 0.5)
```

These functions are built on top of `ggplot2`, so we can add any ggplot layers to our plot.

```{r}
gbinom(100, 0.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  xlim(c(25, 75)) +
  theme_classic() +
  labs(
    title = "Binom(100, 0.5)",
    subtitle = "From Theory",
    y = "Theoretical Probability"
  )


# a and b represent the left and right endpoints of the range you want to graph!
gbinom(100, 0.5) +
  geom_binom_density(n = 100, p = 0.5, a = 45, b = 50, col = "goldenrod2")
```


We can use the `gnorm` function to plot normal bell curves, which is the next type of RV we will discuss.  Here's a normal curve called a standard normal, which is centered at 0 and has a specific amount of area within a distance of 0.

```{r}
gnorm(0, 1)
```

Normal curves are continuous, so it can be useful to shade in areas under the curve as well.

```{r}
gnorm(0, 1) +
  geom_norm_fill(0, 1, b = 1, fill = "dodgerblue")
```










