---
title: 'Section 2: Setup'
author: "Student's Name Here"
date: "2024-09-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Markdown

Markdown is a lightweight markup language for creating pretty formatted text.  In the editor, it appears as plaintext. 

- Can do *italicized* or **bolded** text
- Create headings with the # symbol

# Fewer hashtags = bigger

### More hashtags = smaller

It's best practice to separate different items in Markdown with an empty line.
Regular line break doesn't show up in the formatted file.

Here is the other component of an .Rmd, the R code chunk.

```{r}
# This is a "comment"; preceded by #

# It allows you to write plain text within code chunks, helpful for adding explanations to larger code chunks.

# The next line of code (notice it has no # before it and it is a different color) runs a command, getwd(), which stands for "get working directory", which should be the location of this file.

getwd()
```

---

`TRUE` and `FALSE` are special reserved words. (Or `T` and `F` for short).

```{r}
5 == 15 / 3
3 == 6

T
F
```

Math operations can result in `NaN`, `Inf` or `-Inf` (to avoid breaking R).

```{r}
sqrt(-4)
0 / 0

1 / 0

log(0)
```








