---
title: "week11-in class notes"
output: html_document
date: "2024-11-11"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("~/Desktop/STAT 240/scripts/ggprob.R")
```


# Sec 006 Monday 11/11

```{r}
sum1 = chimpanzee %>% 
  mutate(session_type = case_when(
    partner == "none" ~ "no partner",
    TRUE ~ "partner"
  )) %>% 
  group_by(actor, session_type) %>% 
  summarize(prosocial = sum(prosocial),
            selfish = sum(selfish),
            n = prosocial + selfish,
            pct_prosocial = 100*prosocial/n)
sum1
```

```{r}
visualizeProportionCI = function(true_p, sample_size, conf.level = 0.95, theory = TRUE, wald = FALSE) {
  stdError = sqrt(true_p * (1 - true_p) / sample_size)
  
  # calculate true sampling distribution curve
  samplingDistribution = tibble(
    x = seq(true_p - 3*stdError, true_p + 3*stdError, length.out = 50),
    y = dnorm(x, true_p, stdError)
  )
  
  # calculate random point estimate, associated confidence interval
  moeLabel = paste0("Interval width: This region contains ", 100*conf.level, "% of area")
  point_estimate = rnorm(1, true_p, stdError)
  if (wald) {
    stdError = sqrt(point_estimate * (1 - point_estimate)/sample_size)
    moeLabel = paste0("ESTIMATE: This region (on avg.) contains ", 100*conf.level, "% of area")
  }
  moe = qnorm(1 - (1-conf.level)/2) * stdError
  ciLeft = point_estimate - moe
  ciRight = point_estimate + moe
  
  # encode confidence interval as green if it contains the true parameter, red if it does not
  myColor = ifelse(between(true_p, ciLeft, ciRight), "chartreuse3", "red")
  
  myAlpha = 0.4
  # for more "realistic" graphing, remove the color and the helpful visuals
  if (!theory) {
    myColor = "black"
    myAlpha = 0
  }

  helpfulY = dnorm(true_p, true_p, stdError)/20 # helpful margin to keep things scaled nicely graphically
  moeY = dnorm(true_p + moe, true_p, stdError) # y level of MOE visual
  
  ggplot() +
    # underlay sampling distribution
    geom_line(aes(x, y), samplingDistribution, color = "gray", alpha = 2*myAlpha) +
    geom_hline(yintercept = 0) +
    
    # add true parameter value visual
    geom_vline(xintercept = true_p, linetype = "dashed", color = "blue", size =  1.5, alpha = myAlpha) +
    annotate("text", x = true_p - stdError/5, y = dnorm(true_p, true_p, stdError) - helpfulY, color = "blue", alpha = myAlpha, angle = 90, hjust = "right", vjust = "top", label = "True Parameter", size =  6) +
    
    
    # annotate margin of error visual and label
    annotate("segment", x = true_p - moe, xend = true_p + moe, y = moeY, yend = moeY, arrow = arrow(ends = "both", length = unit(.2, "cm")), alpha = myAlpha) +
    annotate("text", x = true_p - conf.level*moe, y = moeY + helpfulY/2, hjust = "left", vjust = "bottom", label = moeLabel, alpha = myAlpha) +
    
    # annotate point estimate and label
    annotate("segment", x = point_estimate, xend = point_estimate, y = 0, yend = helpfulY*2, size = 2) +
    annotate("text", x = point_estimate, y = -helpfulY, label = "PE", size = 4) +
    
    # annotate confidence interval
    annotate("segment", x = ciLeft, xend = ciRight, y = helpfulY, yend = helpfulY, color = myColor, size = 1.5, arrow = arrow(ends = "both", length = unit(.2, "cm"))) +
    
    annotate("text", x = true_p + stdError, y = dnorm(true_p, true_p, stdError), label = paste0("Sampling Dist: N(", round(true_p, 2), ", ", round(stdError, 2), ")"), alpha = myAlpha) +
    
    theme_minimal() + 
    theme(
      panel.grid = element_blank(),
      axis.text.y = element_blank(),
      axis.text.x = element_text(size = 10)) + 
    
    labs(
      x = "Potential Values of p",
      y = ""
    )
}

```

```{r}
visualizeProportionCI(true_p = 0.4, sample_size = 200, conf.level = 0.95, wald = TRUE)
visualizeProportionCI(true_p = 0.4, sample_size = 200, conf.level = 0.95)
```

# Sec 006 Wednesday 11/13

```{r}
B = 1000000
n = 90
x = 60
p = x/n

## do the simulation
sim = tibble(
  x_star = rbinom(B, n, p),
  p_hat = x_star / n)

## summarize the simulation
sim_summary = sim %>% 
  summarize(mean = mean(p_hat),  ## this should be very close to p=60/90=0.667
            sd = sd(p_hat)) ## this should be very close to sqrt(2/3 * 1/3 / 90) = 0.0497
sim_summary

ggplot(sim, aes(x = p_hat, y=after_stat(density))) +
  geom_histogram(center = 60/90, binwidth = 1/90,
                 color = "black", fill = "firebrick") +
  xlab("sample proportion (p-hat)") +
  theme_minimal() +
  # Theoretical sampling distribution
  geom_norm_density(mu=p, sigma=sqrt(p*(1-p)/n),color="blue", size=2) 
```

## Wald Approximation

```{r}
n <- 90
c = 0.95
p_hat <- 60/90
p_hat
cv <- qnorm(c+(1-c)/2)
c(p_hat - cv*sqrt(p_hat*(1-p_hat)/n), p_hat + cv*sqrt(p_hat*(1-p_hat)/n))
```

## Agresti-Coull Approximation

```{r}
x <- 60
n <- 90
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC - cv*sqrt(p_hat_AC*(1-p_hat_AC)/n), p_hat_AC + cv*sqrt(p_hat_AC*(1-p_hat_AC)/n))
```

Extreme values of p_hat

```{r}
n <- 90
c = 0.95
p_hat <- 0.1
p_hat
cv <- qnorm(c+(1-c)/2)
c(p_hat - cv*sqrt(p_hat*(1-p_hat)/n), p_hat + cv*sqrt(p_hat*(1-p_hat)/n))
```

```{r}
x <- 9
n <- 90
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC - cv*sqrt(p_hat_AC*(1-p_hat_AC)/n), p_hat_AC + cv*sqrt(p_hat_AC*(1-p_hat_AC)/n))
```

## Hypothesis Testing

-   Model Building: $$X \sim Binom(200, p)$$

-   State the hypothesis: $$H_0: p = 0.25, H_{\alpha}: p>0.25$$

-   Identifying the test statistic and null distribution: $$X \sim Binom(200, 0.25)$$

-   Calculate the p-value: Identify $\hat{p} = 57/200$

```{r}
p_value <- 1 - pbinom(56, 200, 0.25)
p_value
```

-   Make conclusion: $\alpha = 0.05$. Since $p-val > \alpha$ we fail to reject H_0

# Sec 003 Wednesday 11/13

```{r}
B = 1000000
n = 90
x = 60
p_hat = x/n

## do the simulation
sim = tibble(
  x_star = rbinom(B, n, p),
  p_hat = x_star / n)

## summarize the simulation
sim_summary = sim %>% 
  summarize(mean = mean(p_hat),  ## this should be very close to p=60/90=0.667
            sd = sd(p_hat)) ## this should be very close to sqrt(2/3 * 1/3 / 90) = 0.0497
sim_summary
```

Build a 95% confidence interval for the chimpanzee making prosocial choices

```{r}
x <- 60
n <- 90
p_hat <- x/n
p_hat
c <- 0.95
cv <- qnorm(c+(1-c)/2)
c(p_hat - cv*sqrt(p_hat*(1-p_hat)/n), p_hat + cv*sqrt(p_hat*(1-p_hat)/n)) # this is using the wald adjustment
```

```{r}
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC-cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)), p_hat_AC+cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)))
```

```{r}
x <- 9
n <- 90
p_hat <- x/n
p_hat
c <- 0.95
cv <- qnorm(c+(1-c)/2)
c(p_hat - cv*sqrt(p_hat*(1-p_hat)/n), p_hat + cv*sqrt(p_hat*(1-p_hat)/n)) # this is using the wald adjustment
```

```{r}
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC-cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)), p_hat_AC+cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)))
```

Hypothesis testing for the Chimpanzee A making prosocial choices

-   Model Statement: X: no of prosocial choice Chimpanzee A makes

BINS: satisfied

$$X \sim Binom(90, p)$$ - Stating the hypothesis:

$$H_0: p = 0.5, H_{\alpha}: p > 0.5 $$ - Check test statistic and figuring out the null distribution

$$X \sim Binom(90, 0.5)$$

-   Calculate the $\hat{p}$

```{r}
p_hat <- x/n
```

-   Calculate the p-value

```{r}
1 - pbinom(59, 90, 0.5)
```

-   Decide a significance level and make your conclusion:

$$\alpha = 0.05$$

Since $p-val < \alpha$, we have enough evidence to reject $H_0$ or support the fact that Chimpanzee A makes prosocial choices.

# Sec 005 Wednesday 11/13

```{r}
B = 1000000
n = 90
x = 60
p = x/n

## do the simulation
sim = tibble(
  x_star = rbinom(B, n, p),
  p_hat = x_star / n)

## summarize the simulation
sim_summary = sim %>% 
  summarize(mean = mean(p_hat),  ## this should be very close to p_hat = 60/90=0.667
            sd = sd(p_hat)) ## this should be very close to sqrt(2/3 * 1/3 / 90) = 0.0497
sim_summary
```

```{r}
ggplot(sim, aes(x = p_hat, y=after_stat(density))) +
  geom_histogram(center = 60/90, binwidth = 1/90,
                 color = "black", fill = "firebrick") +
  xlab("sample proportion (p-hat)") +
  theme_minimal() +
  # Theoretical sampling distribution
  geom_norm_density(mu=p, sigma=sqrt(p*(1-p)/n),color="blue", size=2) 
```

Build a 95% Confidence Interval for the prosocial choice of Chimpqnzee A

### Wald adjustment

```{r}
x <- 60
n <- 90
p_hat <- x/n
p_hat
c <- 0.95
cv <- qnorm(c+(1-c)/2)
c(p_hat-cv*sqrt(p_hat*(1-p_hat)/n), p_hat+cv*sqrt(p_hat*(1-p_hat)/n))
```

###. Agrestti - Coull adjustment

```{r}
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC - cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)), p_hat+cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)))
```

```{r}
x <- 9
n <- 90
p_hat <- x/n
p_hat
c <- 0.95
cv <- qnorm(c+(1-c)/2)
c(p_hat-cv*sqrt(p_hat*(1-p_hat)/n), p_hat+cv*sqrt(p_hat*(1-p_hat)/n))
```

###. Agrestti - Coull adjustment

```{r}
p_hat_AC <- (x+2)/(n+4)
p_hat_AC
c(p_hat_AC - cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)), p_hat_AC + cv*sqrt(p_hat_AC*(1-p_hat_AC)/(n+4)))
```

## Hypothesis testing: To test whether Chimpanzee A actually makes prosocial choices.

-   Model:

$$X_{partner}: \text{no of times Chimpanzee A chooses prosocially}$$ BINS: satisfied

$$X_{partner} \sim Binom(90, p)$$ - State the hypothesis:

$$H_0: p \leq 0.5, H_{\alpha}: p>0.5$$

-   Figuring the test statistic and null distribution:

$$z = \frac{\hat{p}-p_{null}}{s.e(\hat{p})}$$ Null distribution is $X_{partner}\sim Binom(90, 0.5)$

-   Identifying $\hat{p}=0.667$ or X = 60

-   Calculate the p-value

```{r}
1 - pbinom(59, 90, 0.5)
```

-   Decide a significance level $\alpha = 0.05$:

Since $p-val < \alpha$, we reject $H_0$, or we have enough evidence to support that Chimpanzee A makes more prosocial choices.

Since the confidence interval doesn't include any value equal or less than 0.5, we know that it supports the findings of the hypothesis testing.


# Sec 006 Friday 11/15

Hypothesis testing for the proportion of prosocial choices that Chimpanzee A makes when he has a parter (p_partner).

- Step 1: Model Statement: X is the number of prosocial choices made by Chimpanzee A when he has a partner.

`BINS` assumption: satisfied

$$X \sim Binom(90, p)$$

- Step 2: State your hypothesis:

$$H_0: p\leq 0.5 (p = 0.5), H_{\alpha}: p > 0.5$$

- Step 3: Identify your test statistic or null distribution:

$$t = \frac{\hat{p}-p_{null}}{s.e(\hat{p})} \text{ we will not use this statistic and stick with the X}$$

Under the null hypothesis $t = \frac{\hat{p}-0.5}{s.e(\hat{p})} \sim N(0,1)$. 

We can also use the distribution of X, which is $X \sim Binom(90, 0.5)$. 

- Step 4: Calculate the p-value

```{r}
X_sample = 60
n = 90
p_val <- 1 - pbinom(59, 90, 0.5)
p_val
```

- Step 5: Compare your p-val with the significance

$$\alpha = 0.05, 0.02$$

Since, p-val is less than alpha, we reject our null hypothesis. We have enough evidence to support the fact that Chimpanzee makes proscocial choices.

# Confidence interval for difference of proportions

```{r}
x_1 <- 60
n_1 <- 90
p_1_AC <- (x_1+1)/(n_1+2)
p_1_AC
x_2 <- 16
n_2 <- 30
p_2_AC <- (x_2+1)/(n_2+2)
p_2_AC

cv <- qnorm(0.95+(1-0.95)/2)
cv

p.e <- p_1_AC-p_2_AC
s.e <- sqrt((p_1_AC*(1-p_1_AC)/(n_1+2))+(p_2_AC*(1-p_2_AC)/(n_2+2)))
s.e
c(p.e - cv*s.e, p.e + cv*s.e)
```

We are 95% confident that the differnce of proportion lies between (-0.06625432, 0.32984128).





# Sec 003-Friday 11/15


Hypothesis testing for the psychic example

- Model Statement:

$X=\text{no of times the psychic guesses suit correctly}$

`BINS` assumption: satisfied

$$X \sim Binom(200,p)$$

- State the hypothesis:

$$H_0: p=0.5, H_{\alpha}: p > 0.25$$
- Identify the test statistics and the null distribution

$$t = \frac{\hat{p}-p_{null}}{s.e(\hat{p})} \sim N(0,1)$$
We are sticking with $X \sim Binom(200, 0.25)$, 

- Step 4: Calculate the p-value

```{r}
x_sample = 57
p_val <- pbinom(56, 200, 0.25, lower.tail = F)
p_val
```

- Make conclusion depending on the significance level

$$\alpha = 0.05$$
We fail to reject the null hypothesis or we have enough evidence to suggest that the psychic is guessing randomly.


## Difference of proportion

- 95% Confidence interval using the Agresti-Coull

```{r}
x_1 <- 60
p1_hat <- 60/90
p1_hat_AC <- (60+1)/(90+2)
n1 <- 90
x_2<- 16
p2_hat <- 16/30
p2_hat_AC <- (16+1)/(30+2)
n2 <- 30

p.e <- p1_hat_AC - p2_hat_AC
s.e <- sqrt((p1_hat_AC*(1-p1_hat_AC)/(n1+2))+(p2_hat_AC*(1-p2_hat_AC)/(n2+2)))
cv <- qnorm(0.95+(1-0.95)/2)

c(p.e - cv*s.e, p.e + cv*s.e)
```


```{r}
p_bar <- (x_1+x_2)/(n1+n2)
s.e = sqrt((p_bar*(1-p_bar)/n1)+(p_bar*(1-p_bar)/n2))
t = (p1_hat-p2_hat)/s.e
t
```

# Sec 005 Friday 11/15

- Step 1: $X = \text{no. of times the psychic gets the suit of the card right}$

BINS assumption: satisfied

$$X \sim Binom(200, p)$$
- State Hypothesis:

$$H_0: p=0.25, H_{\alpha}: p >0.25 $$
- Identify your test statistic and null distribution:

$$t = \frac{\hat{p} - p_{null}}{s.e(\hat{p})} \sim N(0,1)$$

We can use the distribution of $X$ under $H_0$, $$X_{null} \sim Binom(200, 0.25).$$

- Calculate p-value

```{r}
x_sample = 57
p_val <- 1 - pbinom(56, 200, 0.25)
p_val
```

- Making a conclusion using the significance

$$\alpha = 0.05$$

Since p-val is more than the significance level, we fail to reject the null hypothesis. We have enough evidence that the psychic is guessing at random.



## Difference of proportions

95% Confidence Interval

```{r}
p1_hat <- 60/90
p1_hat_AC <- (60+1)/(90+2)
n1 <- 90
p2_hat <- 16/30
p2_hat_AC <- (16+1)/(30+2)
n2 <- 30

p.e <- p1_hat_AC - p2_hat_AC
s.e <- sqrt((p1_hat_AC*(1-p1_hat_AC)/(n1+2))+(p2_hat_AC*(1-p2_hat_AC)/(n2+2)))
c.v <- qnorm(0.95+(1-0.95)/2)

c(p.e - c.v*s.e, p.e + c.v*s.e)
```

- Hypothesis testing:

```{r}
p_global <- (60+16)/(90+30)

t <- (p1_hat-p2_hat)/(sqrt(p_global*(1-p_global)/90 + p_global*(1-p_global)/30))
t
```

```{r}
p_val <- 1 - pnorm(t)
p_val
```

Considering $\alpha = 0.05, 0.10$, since the p-val is bigger than 0.05 we fail to reject $H_0$. We have enough evidence that the proportion of choices that Chimpanzee A makes with partner or no partner is the same. 

# Sec 003 Monday 11/18

- Step 3: Test statistic and null distribution

$$t = \frac{(\hat{p}_1 - \hat{p}_2) - (p_1-p_2)}{s.e(\hat{p}_1 - \hat{p}_2)}$$
We replace $p_1 \text{ and } p_2$ in the $s.e(\hat{p}_1 - \hat{p}_2)= \sqrt{\frac{p_1(1-p_1)}{n_1}+\frac{p_2(1-p_2)}{n_2}}$ with $\bar{p}= \frac{X_1+X_2}{n_1+n_2}$


```{r}
x_1 <- 60
n_1 <- 90
p1_hat <- (x_1)/(n_1)
p1_hat
x_2 <- 16
n_2 <- 30
p2_hat <- (x_2)/(n_2)
p2_hat

p_bar <- (x_1+x_2)/(n_1+n_2)
p_bar
s.e <- sqrt((p_bar*(1-p_bar)/n_1)+(p_bar*(1-p_bar)/n_2))

t = (p1_hat-p2_hat)/(s.e)
t
```

$$H_0: p_1 = p_2, H_{\alpha}: p_1>p_2$$

```{r}
p_val <- 1 - pnorm(t)
p_val
```

# Sec003 Monday 11/18

```{r}
x1 <- 60
n1 <- 90
x2 <- 16
n2 <- 30

p1_hat <- x1/n1
p2_hat <- x2/n2

p_bar <- (x1+x2)/(n1+n2)
p_bar
s.e <- sqrt((p_bar*(1-p_bar)/n1) + (p_bar*(1-p_bar)/n2))
s.e
t <- (p1_hat - p2_hat)/s.e
t

p_val <- 1 - pnorm(t)
p_val
```



```{r}
x_bar <- 235.5062
s <- 36.2
n <- 3557

s.e <- s/sqrt(n)
s.e

cv <- qt(0.975, n-1)
cv

#qnorm(0.975)

x_bar + c(-1,1)*cv*s.e
```


# Sec005 Monday 11/18

95% Confidence Interval: 

$$(\bar{x} - qt(c+(1-c)/2, n-1)*\frac{s}{\sqrt{n}}, \bar{x} - qt(c+(1-c)/2, n-1)*\frac{s}{\sqrt{n}})$$

```{r}
n <- 3557
x_bar <- 235.5062
s <- 36.2

s.e <- s/sqrt(n)
s.e
cv <- qt(0.975, n-1)
cv

# qnorm(0.975)

x_bar + c(-1,1)*cv*s.e
```










