---
title: "Homework 11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
source("../../scripts/ggprob.R")
```

# Preliminaries

- This file should be in `STAT240/homework/hw11` on your local computer.
- Download `parental_leave.csv` to `STAT240/data`.


# Problem 1

The file `parental_leave.csv` contains information on crowdsourced parental leave data for 1601 different companies.  

```{r}
leave <- read_csv("../../data/parental_leave.csv")

# Remove spaces from column names
names(leave) <- make.names(names(leave))

# Tidy numeric columns
leave <- leave %>%
  mutate_at(vars(Paid.Maternity.Leave, Unpaid.Maternity.Leave,
                 Paid.Paternity.Leave, Unpaid.Paternity.Leave), as.numeric) %>%
  replace(is.na(.), 0)
```

(a) Mutate a column for total paid leave by adding the hours of paid maternity and paternity leave.

- Optionally, give the columns shorter names that are easier to reference.

```{r}
# Write your code here!
```

(b) Filter the data to only include the "Educational Services: College & Universities" industry.  Create a histogram of total paid leave and comment on its shape.

```{r}
# Write your code here!
```

> Replace this text with your response!

(c) Comment on why a T test for the mean might still be an appropriate method, even though the total paid leave data does not look normal.

> Replace this text with your response!

# Problem 2

Build a 98% T CI for the average total paid leave of companies in "Educational Services: College & Universities", and interpret the results.

```{r}
# Write your code here!
```

> Replace this text with your response!


# Problem 3

Perform a test at the 2% level to determine whether the "Educational Services: College & Universities" average total paid leave is greater than 10 weeks.  State hypotheses, report a test statistic and p-value. You must do this by hand, then check your results with `t.test`.

```{r}
# Write your code here!
```

> Replace this text with your response!

# Problem 4

Re-calculate the p-value from problem 3, but use a standard normal distribution instead of a $T_{n-1}$ distribution.  Why are the results very similar?

```{r}
# Write your code here!
```

> Replace this text with your response!

# Problem 5

Now, consider comparing companies in "Educational Services: College & Universities" (ESCU) to companies in "Healthcare: Hospitals & Clinics" (HHC).  Build a 95% CI for the difference in average total paid leave and interpret the results.

```{r}
# Write your code here!
```

> Replace this text with your response!


# Problem 6

Does the educational industry give more paid leave than the healthcare industry?  Perform a hypothesis test of 
$$H_0: \mu_{ESCU} - \mu_{HHC} = 0 \quad \text{versus}\quad H_A: \mu_{ESCU} - \mu_{HHC} > 0$$
with $\alpha = 0.05$.  Do this by hand, then check your results with `t.test`.

```{r}
# Write your code here!
```

# Problem 7

Repeat the test in problem 6, but instead write hypotheses for $\mu_{HHC} - \mu_{ESCU}$ (the order of subtraction is switched).  Show that you get identical results to problem 5.

```{r}
# Write your code here!
```


# Problem 8

Now, consider the "Technology: Software" industry.  Perform the appropriate hypothesis test to determine whether there is a difference in paid maternity and paternity leave within this industry. Note that each company decides its own policies on maternity and paternity leave, which should NOT be considered independent.

```{r}
# Write your code here!
```

> Replace this text with your response!













