---
title: "Section 7: Joining and Pivoting"
author: "Student's Name Here"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```


## Mutating joins

Let's create some example dataframes.

```{r}
drinks <- tibble(
  Row = 1:4,
  Drink = c("Milk Tea", "Espresso", "Latte", "Espresso"),
  Cafe = c("Happy Tea", "Daily Grind", "Daily Grind", "Bev's Beverages"),
  Price = c(4, 2, 3, 2)
)

discounts <- tibble(
  Row = 1:3,
  Coupon = c("25%", "50%", "5%"),
  Cafe = c("Happy Tea", "Daily Grind", "The Roast")
)

drinks
discounts
```

### Left and right joins

`left_join` will merge the datasets and keep all values in the leftmost (x) dataframe.  We specify the column to join by.

```{r}
left_join(x = drinks, y = discounts, by = join_by(Cafe))

# Pipe notation
drinks %>% 
  left_join(discounts, join_by(Cafe))
```

`right_join` does the opposite, by keeping all values in the rightmost (y) dataframe.  Its functionality is identical to `left_join`.

```{r}
discounts %>% 
  right_join(drinks, join_by(Cafe))
```


### Different column names

In the original discounts and drinks data, both have a column named `Cafe`, but the column names do not always match up.

```{r}
discounts2 <- discounts %>% 
  rename(Location = Cafe)

discounts2
```

What happens if we `right_join` without specifying `by`?

```{r}
right_join(discounts2, drinks)
```

It matched `Row` since it shares a name in both x and y.  If the dataframes have no rows in common, we get an error.

```{r}
discounts2 %>% 
  rename(Discount_Row = Row) %>% 
  right_join(drinks)
```

Specify the correct columns in `by`, in the same order as the datasets.

```{r}
discounts2 %>% 
  right_join(drinks, join_by(Location == Cafe))
```


### Full join

`full_join` will keep all rows in both x and y.  Imagine doing a right or left join, and just appending any extra, non-matching rows at the bottom.

```{r}
full_join(drinks, discounts, join_by(Cafe))
```

This is different from simply stacking the rows.

```{r}
bind_rows(drinks, discounts)
```


### Inner join

`inner_join` is the most "selective".  It only keeps rows that appear in both datasets and it does not induce any NA values.

```{r}
drinks %>% 
  inner_join(discounts, join_by(Cafe))
```


### Predict the join result

Guess the number of rows and columns that will result from the join.  Also, will any NA values appear?  Uncomment the join to see if you were correct.

```{r}
band_instruments

band_members
```

```{r}
#left_join(band_instruments, band_members, join_by(name))

#full_join(band_instruments, band_members, join_by(name))

# I did not specify a join_by(). Which column(s) will it match by default?

#inner_join(band_instruments, band_members)

```


## Filtering joins

Filtering joins decide to keep or remove rows, similar to `filter`.  `semi_join` returns matching rows.

What drinks do I have a discount for?  The order matters here.

```{r}
drinks %>% 
  semi_join(discounts, join_by(Cafe))

discounts %>% 
  semi_join(drinks, join_by(Cafe))
```

Notice the differences with `inner_join`.  `inner_join` results in a new column and `semi_join` does not.

```{r}
drinks %>% 
  inner_join(discounts, join_by(Cafe))
```


Use `anti_join` to filter by non-matching rows.  What drinks do I not have a coupon for?

```{r}
drinks %>% 
  anti_join(discounts, join_by(Cafe))
```

Which coupons do not apply to the cafes I go to?

```{r}
discounts %>% 
  anti_join(drinks, join_by(Cafe))
```


### Predict the join result

Revisit the `band_members` and `band_instruments` data.  Predict the output of the following filtering joins.

```{r}
band_members

band_instruments
```

```{r}
#semi_join(band_members, band_instruments, join_by(name))

# Switched order of band_instruments and band_members
#semi_join(band_instruments, band_members, join_by(name))

#anti_join(band_members, band_instruments, join_by(name))
```


## Pivoting

The `players` dataset is in wide form.  There is one row per player, and columns for different statistics.

```{r}
players <- tibble(
  player = c("Erling Haaland", "Kevin De Bruyne"),
  goals = c(36, 7),
  assists = c(8, 16),
  games = c(35, 32)
)

players
```

This table shows the same data in long form.  Now, there is one row per player-statistic.  There are more rows and fewer columns.

```{r}
tibble(
  player = c(rep("Erling Haaland", 3), rep("Kevin De Bruyne", 3)),
  statistic = rep(c("goals", "assists", "games"), 2),
  value = c(36, 8, 35, 7, 16, 32))
```

These two sets of data have the exact same information, organized in different ways.  Pivoting lets us transform the data between long and wide forms.

### pivot_longer

`pivot_longer` is used to turn a wide dataframe into a long dataframe.  We provide the columns that we want to merge into a single column.

```{r}
players

players %>% 
  pivot_longer(c(goals, assists, games))

# Or, specify that we only want to keep "player"
players %>% 
  pivot_longer(-player)

# Can also specify a range of columns
players %>% 
  pivot_longer(goals:games)
```

The columns are automatically called "name" and "value".  We can give these columns are own names.

```{r}
players %>% 
  pivot_longer(-player, names_to = "type", values_to = "amount")
```


### pivot_wider

`pivot_wider` is used to turn a long dataframe into a wide dataframe.  Let's re-do the pivot we did with the players data.

```{r}
players_long <- players %>% 
  pivot_longer(c(goals, assists, games), names_to = "statistic")

players_long
```

In our case, we want to split up the `name` column into three separate columns.  The new column names (goals, assists, games) come from the values of `name`.  We will populate these columns with the data in the `value` column.

```{r}
players_long %>% 
  pivot_wider(names_from = statistic, values_from = value)
```


### Pivoting practice

`table1` is a built-in dataframe of tuberculosis cases by country.

```{r}
table1
```

This dataset is in wide form.  Use `pivot_longer` to convert the table to long-form.  

- There should be one row per country-year-type combination
- There should be four columns: `country`, `year`, `type`, `count`
- `type` specifies whether we are looking at total population or TB cases
- `count` gives the number of people

```{r}
table1 %>%
  pivot_longer()
```


`nests` below is an example dataset of eagle nests with region, year (either 2007 or 2009), and number of nests.

```{r}
nests <- tibble(
  region = c(rep("Pacific", 2), rep("Southwest", 2), rep("Rocky Mts", 2)),
  year = rep(c(2007, 2009), 3),
  num_nests = c(1039, 2587, 51, 176, 200, 338)
)

nests
```

This dataset is in long form. Use `pivot_wider` to convert the table to wide form.

- There should be one row per year
- There should be four columns: `year`, `Pacific`, `Southwest`, `Rocky Mts`
- The latter three columns give the number of nests in that region


```{r}
nests %>%
  pivot_wider()
```





