---
title: "Discussion 6"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/discussion/ds06`.
- Please also download `Messi_Through_Oct112024_FBRef.csv` to `STAT240/data`.

- This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Member 1

> Present Group Member 2

> Present Group Member 3

> Present Group Member 4 (some groups only have 3)

# Problem 1

The following dataset contains information on all the games Lionel Messi has played in the 2024 season, sourced from [FBRef.com](https://fbref.com/en/players/d70ce98e/matchlogs/2024/Lionel-Messi-Match-Logs).

```{r}
messi_original = read_csv("../../data/Messi_Through_Oct112024_FBRef.csv")
```

Each row is a single game. The `Gls` column indicates how many goals Messi scored in that game, and `Ast` indicates how many assists Messi provided in that game. (An assist is a pass that leads to a goal.)

**Create and save a new dataframe** called whatever you want, by executing the following steps, starting from `messi_original`, 

- Create a new column called `points`, calculated as `Gls + Ast`. (You can think of this as how many goals Messi was involved in during that game - a term borrowed from hockey.)

- Summarize to obtain one row for each unique value of `points`, with a new column `n` indicating in how many games Messi scored that exact value of `Points`.
*Check: Your dataframe should indicate that Messi has scored 0 points 10 times, 1 point 8 times... and one time only he scored 6 points!*

- Create a new column called `percent`, which indicates in what percent of games Messi scored that exact value of points. *(Hint: The total number of games played can be figured out exclusively from the `n` column!)*

```{r}
# Write your code here!

```

# Problem 2

`points` and `percent` from the dataframe above form a **probability distribution**.

Is it a **discrete** probability distribution or a **continuous** one? Briefly explain why.

> Replace this text with your response.


# Problem 3

Using your dataframe from problem 1, what is the **average number of points** Messi has scored in games in 2024?

```{r}
# Write your code here!

```

# Problem 4

**Plot** the probability distribution, with a vertical bar for each value of `points`, the height of which represents `percent`.

Incorporate the following customizations:

- Adjust the X axis so that the ticks and labels indicate every whole number 0 through 6.

- Add a **red, vertical, dashed** line at the mean you calculated in problem 3.

- Make at least one other fun visual improvement that you think makes the graph look better - adjust the colors, add a theme, et cetera!

```{r}
# Write your code here!

```

# Problem 5

For context, Messi made his professional soccer debut in 2004 at the age of 17, and is now 37 and still playing.

The average you calculated in problem 3 and graphed in problem 4 should have been about 1.27.

**Is it valid to say that in a random game from Messi's entire career, we would expect him to score about 1.27 points?** If not, why not?

> Replace this text with your response.

# Problem 6

Using your dataframe from problem 1, what is the **variance** and **standard deviaton** of the number of points Messi has scored in 2024?

```{r}
# Write your code here!

```

# Problem 7

In what percent of 2024 games did Messi score multiple points?

*This would be easy to do by hand; but try to do it by code using your dataframe from problem 1!*

```{r}
# Write your code here!

```

# Problem 8

The percentage you calculated above should be 30.8%. 
**a)**
Is it valid to say that in a randomly selected game from Messi's entire career, there is a 30.8% chance he scored multiple points?

> Replace this text with your response.

Is it valid to say that in a randomly selected game Messi played in 2024, there is a 30.8% chance he scored multiple points?

> Replace this text with your response.


