---
title: "Discussion 9"
output: html_document
---

# Preliminaries

-   This file should be in `STAT240/discussion/ds09` on your local computer.

-   This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

- Download `boston-marathon.csv` and the `diabetes.csv` data sets to `STAT240/data`

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Members:

> Absent Group Members:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
source("../../scripts/ggprob.R")
```


# Problem 1

In a psychological experiment, a reported psychic tries to guess the suit of the cards of a well-shuffled stack of cards which includes many decks shuffled together.

You may assume that the suit of each card in order is equally likely to be any one of the four suits and that the suits of the cards are independent of each other.

**(a)** If $X$ is the number of cards correctly guessed, what distribution does X follow? *Note: You will only know the value of one of the two parameters!*

> Write your answer here!

In guesses of the suits of 100 cards, the psychic correctly guesses the suit 29 times. Conduct a formal hypothesis test to examine the evidence that the psychic can guess the suit of the cards more often than expected, e.g. better than random chance.

**(b)** State hypotheses for conducting this test. Also mention the type of the hypothesis test.

> Write your answer here!

**(c)** Calculate the p-value for the above test. *Reminder: The p-value is the probability on the null distribution observing an outcome more extreme than what we actually did! "More extreme" is defined by the alternative hypothesis.*

```{r}
# Write your code here!
```

**(d)** Considering that the significance level is $0.05$, what conclusion can you come to from the above p-value?

> Write your answer here!

# Problem 2

Suppose you are playing a coin flipping game with a friend, where you suspect the coin your friend provided is not a fair coin. In fact, you think the probability the coin lands heads is less than 0.5. To test this, you flip the coin 100 times and observe the coin lands heads 35 times.

**(a)** State the appropriate null and alternate hypotheses for conducting this test. Also mention the type of the hypothesis test.

> Write your answer here!

**(b)** Calculate the p-value for the above test.

> Write your answer here!

```{r}
# Write your code here!
```

**(c)** Considering that the significance level is $0.05$, what conclusion can you come to from the above p-value?

> Write your answer here!

# Problem 3

**(a)** We conduct the 100-coin-flip experiment five times (with the same coin from above, which may or may not be fair) and get values (35, 35, 30, 33, 31). What is the sample mean? 

```{r}
# Write your code here!
```

**(b)** Assuming that the probability of getting heads while flipping the possibly-unfair coin from problem 2 is 0.3. What will be the distribution of the sample means for part (a)?

> Write your answer here!

**(c)** What is the probability of getting a sample mean smaller than the sample mean in part (a) in the distribution from part (b)?

```{r}
# Write your code here!
```



# EXAM 2 CONTENT STOPS HERE. Any content below will not be on Exam 2.

# Problem 4

Given the value of correlation what relationship does X and Y have? 

- $r = -0.68$

> Write your answer here!

- $r = -0.89$

> Write your answer here!

- $r = -0.015$

> Write your answer here!

- $r = 0.32$

> Write your answer here!

- $r = 0.97$

> Write your answer here!

- $r = -1$

> Write your answer here!


# Problem 4

Using the Boston marathon dataset

```{r}
# This code is provided for you. Do not change it.
set.seed(70)
marathon <- read_csv("../../data/boston-marathon.csv") %>% filter(Year == 2011) %>% select(Age, Time) %>% slice_sample(n = 500)
```

**(a)** Identify the dependent (y) and independent (x) variables in the Boston Marathon dataset. Calculate the correlation between x and y.  Comment on the strength and magnitude of the linear relationship.

```{r}
# Write your code here!
```

> Write your answer here!

**(b)** Create a scatterplot of x vs y (identified above) that also shows the linear regression line.

```{r}
# Write your code here!
```

**(c)** Build a linear model called `model_marathon` for the Boston marathon dataset and display it's summary.

```{r}
# Write your code here!
```

**(d)** The residual plot for the linear regression model created above is given below and check whether the assumptions of the linear regression model is satisfied for this model.

```{r}
# This code is provided for you. Uncomment it to use it.

#residual = resid(model_marathon)

#ggplot(mapping = aes(x = marathon$Age, y = residual)) +  geom_point() + geom_hline(yintercept = 0) + geom_smooth(se= F)
```


> Write your answer here!


# Problem 5

Consider the diabetes data set.

```{r}
# This code is provided for you. Do not change it.

diabetes <- read_csv("../../data/diabetes.csv") %>% rename(glucose = "Glucose") %>% filter(BMI > 1 & BMI < 50 & !is.na(BMI) & !is.na(glucose) & glucose > 0) %>% select(glucose, BMI)
```

```{r}
# This code is provided for you. Uncomment it to use it.

#ggplot(diabetes, aes(x = BMI, y = glucose)) + geom_point() + geom_smooth(method = "lm", se = FALSE)
```


**(a)** Calculate the correlation between Glucose levels (y variable) and BMI (x variable). Comment on the strength and magnitude of the linear relationship based on the correlation and the above plot. Do you think a linear model seems to be appropriate?


```{r}
# Write your code here!
```


> Write your answer here!


**(b)** Calculate the slope and intercept of a least-squares linear regression model using the formula discussed in class. Check your answers with intercepts produced from `lm()`. After you find the estimates write the expression of the linear regression model (expression for $\hat{Y}_i$).


```{r}
# Write your code here!
```

> Write your answer here!

**(c)** Write a code similar to Problem 4 part (d) and perform a residual analysis of the diabetes linear model to assess the fit. Based on the scatterplot of residuals and comment on the three assumptions:

- Linearity
- Normality
- Constant variance

```{r}
# Write your code here!
```

> Write your answer here!


# Problem 6

**(a)** Depending on the model built in Problem 5, part b, predict the glucose level of a person with a BMI of 51.

```{r}
# Write your code here!
```

**(e)** What conclusions can you make about the likelihood of a person developing diabetes based on their BMI.

> Write your answer here!


