---
output: html_document
---

```{r setup, include = FALSE}
##   when the tidyverse library is loaded, messages are generated that we do not want to include in the knitted document
## setting the chunk argument include = FALSE means this chunk is executed, but the text in the chunk and any output do not appear in the knitted document

## The following command sets some options for this R Markdown file
  ##   echo = TRUE means that the code in R chunks will appear in the knitted document
  ##   message = FALSE and warning = FALSE mean that messages and warnings do not appear in the final document
  ##   error = TRUE means the document will continue to knit in the presence of errors while knitting
  ##   fig.height = 3 means figures are 3 inches tall unless otherwise specified
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE,
                      error = TRUE, fig.height = 3,
                      fig.width = 4)

library(tidyverse)
```


### Grocery Example

```{r}
grocery_prices = read_csv("../../data/grocery-prices.csv")
grocery_list = read_csv("../../data/grocery-list.csv")

produce_orig = grocery_list %>% 
  left_join(grocery_prices, by = "item") 
```

### Column changers

#### mutate()

```{r}
produce = produce_orig %>% 
    mutate(cost = price * quantity,
           sales_tax = 0.05 * cost)
```


#### select()

```{r}
produce %>%
  select(cost, sales_tax)

produce %>%
  select(-quantity)

produce %>% 
  select(starts_with("q"))

produce %>% 
  select(1:4)
```

#### relocate()

```{r}
produce %>% 
  relocate(price, quantity, cost, sales_tax)

produce %>% 
  select(price, quantity, cost, sales_tax)

produce %>% 
  relocate(cost, .after = item)


```



#### rename()

```{r}
produce %>% 
  rename(pricePerUnit = price,
         totalCost = cost)
```


### Row changers

#### arrange()

```{r}
produce %>% 
  arrange(cost)

produce %>%
  arrange(desc(item))

produce %>% 
  arrange(type, price)


```

#### filter()

```{r}
produce %>% 
  filter(type == "fruit")

produce %>% 
  filter(cost > 2 & cost < 5)

produce %>% 
  filter(between(cost, 2, 5))

produce %>% 
  filter(cost < 2 | cost > 5)

produce %>% 
  filter(cost < 2 & cost > 5)
```

#### Exercises

“Show me just the fruits.”

```{r}
produce %>% 
  filter(type == "fruit")
```

“Can you put whether it’s a fruit or vegetable at the beginning?”

```{r}
produce %>% 
  relocate(type)
```

“I only care about the food name and the total cost, can you get rid of everything else?”

```{r}
produce %>% 
  select(item, cost)
```

“Can you put the foods in alphabetical order so apple is first?”

```{r}
produce %>% 
  arrange(item)
```

“Can you give each food a sale price that’s half off the normal price?”

```{r}
produce %>% 
  mutate(sales_price = price / 2) %>% 
  relocate(sales_price, .after = price)

produce %>% 
  mutate(sales_price = price / 2, .after = price)
```

### summarize()

```{r}
produce %>% 
  summarize(totalCost = sum(cost),
            totalTax = sum(sales_tax)) #%>% 
  #as.data.frame()
```

### group_by()

```{r}
produce %>% 
  group_by(type)
```

- Find total cost of fruits and vegetables separately

```{r}
produce %>% 
  group_by(type) %>% 
  summarize(totalCost = sum(cost))
```

- group_by() and mutate() example

Find the % of the cost for each item within the total cost for the type (fruit or vegetable)

```{r}
prob = produce %>% 
  group_by(type) %>% 
  mutate(totalCost = sum(cost), .after = cost) %>% 
  mutate(pct_cost = 100 *cost / totalCost)

prob %>% 
  print(width = Inf)
  
```

```{r}
prob2 = produce %>% 
  group_by(type) %>% 
  mutate(totalCost = sum(cost), .after = cost) %>% 
  mutate(pct_cost = 100 *cost / totalCost) %>% 
  select(item, type, cost, totalCost, pct_cost) #%>% 
  #ungroup()

prob2

### grouping is still present
prob2 %>% 
  summarize(sum = sum(pct_cost))

prob2 %>% 
  ungroup() %>% 
  summarize(sum = sum(pct_cost))
```

### n()

```{r}
produce %>% 
  summarize(n = n())

produce %>% 
  group_by(type) %>% 
  summarize(n = n())

produce %>% 
  group_by(type) %>% 
  summarize(n = n(),
            total_cost = sum(cost),
            max_price = max(price),
            median_tax = median(sales_tax))
```

### count()

```{r}
produce %>% 
  count(type)

produce %>% 
  count()
```

### slice

```{r}
produce %>% 
  slice_max(quantity)
```

