---
title: "Section 12: Linear Regression"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```

## Correlation

A scatterplot shows a downward trend between year and freeze duration.

```{r}
monona <- read_csv("../../data/lake-monona-winters-2024.csv")

ggplot(monona, aes(x = year1, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm", se = F)
```

We can calculate correlation with the `cor` function.

```{r}
x <- monona$year1
y <- monona$duration

cor(x, y)
cor(y, x)
```


### Correlation examples

```{r}
# Hide this chunk - just for the purpose of making example plots

set.seed(2023)
x1 = seq(-2,2,0.05)
y0 = rnorm(length(x1))
y1 = x1^2
x2 = x1[x1>0]
y2 = y1[x1>0]
x3 = x1[x1<0]
y3 = y1[x1<0]
x4 = x1
y4 = x4 + rnorm(length(x4),0,0.3)
x5 = x1
y5 = x5 + rnorm(length(x5),0,1)
x6 = x1[x1 > -1]
y6 = exp(3*x6)/10 + rnorm(length(x6), 0, 5)

cor_plot = function(x,y)
{
  ggplot(tibble(x,y), aes(x=x, y=y)) +
    geom_point() +
    geom_smooth(se = FALSE, method = "lm") +
    geom_vline(xintercept = mean(x), color = "red", linetype = "dashed") +
    geom_hline(yintercept = mean(y), color = "red", linetype = "dashed") +
    ggtitle(str_c("r ≈ ", round(cor(x,y),2))) +
  theme_bw() +
  theme(text = element_text(size = 20)) 
}
```

Correlation 0 corresponds to no linear relationship.  This can be a horizontal line or some other shape.

```{r}
cor_plot(x1, y0)

cor_plot(x1, y1)
```

Correlation near 1 or -1 indicates that a line fits the data well.  This does not necessarily imply a true linear relationship.

```{r}
cor_plot(x4, y4)
cor_plot(-x4, y4)

cor_plot(x2, y2)
```

A correlation of 0.7 indicates that a line fits the data somewhat well.  Again, there maybe some other underlying relationship.

```{r}
cor_plot(x5, y5)

cor_plot(x6, y6)
```



## Linear modeling

What is the relationship between a baseball team's winning percentage and the latitude of its stadium?

```{r}
set.seed(20240406)
fake_latitude <- tibble(
  win_perc = runif(25, 0, 1),
  latitude = runif(25, 25, 48)
)

ggplot(fake_latitude, aes(latitude, win_perc)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Winning Percentage vs. Latitude of Home Stadium",
       subtitle = "(Not real data)",
       y = "Winning Percentage",
       x = "Latitude (degrees)")
```

Now, consider predicting the winning percentage based on average wages.

```{r}
set.seed(20240406)
fake_wages <- tibble(
  wages = rnorm(25, 5, 2),
  win_perc = wages* rnorm(25, 1, 0.5),
  win_perc_adj = (win_perc - min(win_perc))/max(win_perc - min(win_perc))
)

ggplot(fake_wages, aes(wages, win_perc_adj)) +
  geom_point() +
  geom_smooth(method = "lm", se = F) +
  labs(title = "Winning Percentage vs. Wages",
       subtitle = "(Not real data)",
       y = "Winning Percentage",
       x = "Wages (Millions of $")
```


### Exercise

```{r}
riley <- read_table("../../data/riley.txt")
riley_2_8 <- riley %>% 
  filter(age >= 2*12 & age <= 8*12)

y <- riley_2_8$height
x <- riley_2_8$age

# Find estimates for beta_0 and beta_1
# You'll need to find correlation and mean/sd of x and y

```

We can fit a linear model automatically with `lm`.

```{r}
lm(height ~ age, data = riley_2_8)

# It's helpful to save the result to reference it later
height_mod <- lm(height ~ age, data = riley_2_8)

summary(height_mod)

ggplot(riley_2_8, aes(x = age, y = height)) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm", fullrange = T) +
  ylab("Height (inches)") +
  xlab("Age (months)")
```

We can also do this for our Lake Monona data.

```{r}
lake_mod <- lm(duration ~ year1, data = monona)
```


## Assumptions

We evaluate the assumptions (i.e., validate the linear model fit) by studying the residuals.  For ease of graphing, we can add the residuals to our data.

```{r}
resid(lake_mod)

monona %>% 
  mutate(residuals = resid(lake_mod)) %>% 
  select(year1, duration, residuals) %>% 
  head()
```


### Linearity

```{r}
# This is just a helpful function so I don't have to type this out every time;
# feel free to use it yourself or not as you like

resid_plot <- function(data) {
  # assumes the columns are named "x" and "y" respectively
  data %>%
    mutate(resid = resid(lm(y~x, data))) %>% 
    ggplot(aes(x, resid)) +
    geom_point() +
    geom_hline(yintercept = 0)
}
```

There should be no obvious curved patterns in a residual plot.

```{r}
set.seed(2024040)
linear_data <- tibble(
  x = runif(100, 0, 10),
  y = 0.05 * (x - 5) + rnorm(100, sd = 0.5)
)

nonlinear_data <- tibble(
  x = runif(100, 0, 10),
  y = 0.05 * (x - 5)^2 + rnorm(100, sd = 0.5)
)
```

```{r}
resid_plot(nonlinear_data) +
  ggtitle("Linearity Violated: Curved Pattern Present")

resid_plot(linear_data) +
  ggtitle("Linearity Satisfied: No Curved Pattern Present")
```


### Normality

The points in a residual plot should be evenly scattered around the center horizontal line at 0.

```{r}
set.seed(20240407)
nonnormal_data <- tibble(
  x = runif(100, 0, 10),
  y = 0.05 * (x - 5) + sample(c(-2, 2), 100, replace = TRUE) * rnorm(100, mean = 1, sd = 0.5)
)

set.seed(20240407)
nonnormal_data2 <- tibble(
  x = runif(100, 0, 10),
  y = 2*x + rgamma(100, 1) - 1
)
```

```{r}
resid_plot(nonnormal_data) +
  ggtitle("Normality Violated: Points Don't Tend Towards Center")

resid_plot(nonnormal_data2) +
  ggtitle("Normality Around 0 Violated: Non-Symmetric")

resid_plot(linear_data) +
  ggtitle("Normality Satisfied: Residuals Symmetric Around 0")
```


### Constant variance

The residual plot should show no relationship between x and the spread of residuals.  We don't want to see a "megaphone" shape.

```{r}
set.seed(20240407)
heteroskedastic_data <- tibble(
  x = runif(100, 0, 10),
  y = x + rnorm(100, sd = 0.5 * x)
)

heteroskedastic_data2 <- tibble(
  x = runif(100, 0, 10),
  y = x + rnorm(100, sd = 0.5 * (10-x))
)

```


```{r}
resid_plot(heteroskedastic_data) +
  ggtitle("Constant Variance Violated: Spread Increases with X")

resid_plot(heteroskedastic_data2) +
  ggtitle("Constant Variance Violated: Spread Decreases with X")

resid_plot(linear_data) +
  ggtitle("Constant Variance Satisfied: Equal Spread Across X")
```


### Monona data

Let's verify the fit of the linear model on the lake Monona data.

```{r}
monona %>% 
  mutate(residuals = resid(lake_mod)) %>% 
ggplot(aes(x = year1, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0)
```


## Prediction

How tall was Riley at age 78 months?

```{r}
30.25 + (0.25 * 78)

# or

predict(height_mod, newdata = tibble(age = 78))
```

Graphically:

```{r}
new_x <- 78
new_y  <- 30.25 + (0.25 * new_x)

ggplot(riley_2_8, aes(x = age, y = height)) +
  geom_point() +
  # fullrange = T extends the blue line past the X range of the existing points
  geom_smooth(se = FALSE, method = "lm", fullrange = T) +
  geom_point(aes(x = new_x, y = new_y), color="red", size=4) + 
  geom_vline(xintercept = new_x, color="red", linetype="dashed") +
  geom_hline(yintercept = new_y, color = "red", linetype = "dashed") +
  ylab("Height (inches)") +
  xlab("Age (months)")
```


### Monona data

According to the linear model, what will be the duration of the 2026-2027 freeze? (year1 = 2026)

```{r}
lake_mod <- lm(duration ~ year1, data = monona)

```


We have to be mindful when doing prediction.  Extrapolation can be dangerous!

```{r}
# How long will the freeze in 2400 be?

535.125 + -0.223 * 2400
```



