---
title: "Section 17: Inference for Two Means"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
source("../../scripts/ggprob.R")
```

## CI for a difference in means

### Data tidying

`TIM.txt` contains data on the times of runners of the Boston Marathon from 2010 and 2011 (and 2013, which we will not analyze).

```{r}
bm_orig <- read_table("../../data/TIM.txt",
                col_types = cols(.default = col_double()))  # Columns are 'double' by default
```

We will make the following changes:

- Eliminate data from 2013
- Create a `Sex` variable with better labels
- Sum the times for racers for different race segments to obtain a total time
- Add an `Age_Range` variable matching the competitor classses in the Boston Marathon.

```{r}
bm <- bm_orig %>% 
  filter(!is.na(`K40-Fin`)) %>% 
  filter(Year < 2013) %>% 
  arrange(Year, BibNum) %>% 
  select(-starts_with("Start"), -HalfMar, -Age2014) %>% 
  
  mutate(Sex = case_when(
    Gender1F2M == 1 ~ "female",
    Gender1F2M == 2 ~ "male")) %>% 
  
  mutate(Time = pmap_dbl(select(., starts_with("K")), sum)) %>% 
  
  mutate(age1 = case_when(
    Age < 35 ~ 18,
    Age >= 80 ~ 80,
    TRUE ~ floor(Age/5)*5),
    age2 = case_when(
      Age < 35 ~ 34,
      TRUE ~ age1 + 4),
    Age_Range = case_when(
      age1 == 80 ~ "80 and older",
      TRUE ~ str_c(age1,"-",age2))) %>% 
  
  select(-Gender1F2M, -age1, -age2) %>% 
  relocate(BibNum, Year, Sex, Age, Age_Range, Time)
```

Each row is a single "run" of the marathon.  If a person competed in 2010 and 2011, they would have two rows.  Columns:

- `BibNum`: bib number, a unique identifier for each competitor within a year
- `Year`: either 2010 or 2011
- `Sex`: female or male
- `Age`: age in years
- `Age_Range`: from 18-34, 35-39, ..., 75-79, and 80 and older
- `Time`: total time to finish in decimal minutes
- `K0-5` through `K40-Fin`: times in decimal minutes for different segments of the race


### Data exploration

Let's compare the basic summaries for each group (2010 and 2011).

```{r}
bm_twomeans_summary <- bm %>% 
  filter(Sex == "female", Age_Range == "18-34") %>% 
  group_by(Year) %>% 
  summarize(avgTime = mean(Time), sdTime = sd(Time), n = n())

bm_twomeans_summary
```

Let's also extract the values and save them for later.

```{r}
xbar1 <- bm_twomeans_summary$avgTime[1]
xbar2 <- bm_twomeans_summary$avgTime[2]
s1 <- bm_twomeans_summary$sdTime[1]
s2 <- bm_twomeans_summary$sdTime[2]
n1 <- bm_twomeans_summary$n[1]
n2 <- bm_twomeans_summary$n[2]
```


Let's calculate the approximate df for comparing the difference in means.

```{r}
w_numer <- (s1^2/n1 + s2^2/n2)^2
w_denom <- (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))

w <- w_numer / w_denom
```

Now, we can build a CI

```{r}
pt_est <- xbar1 - xbar2
se <- sqrt(s1^2/n1 + s2^2/n2)

cv <- qt(0.975, df = w)

c(pt_est - cv*se, pt_est + cv*se)
```

Now, confirm with `t.test`.

```{r}
times_2010 <- bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2010)

times_2011 <- bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2011)

t.test(times_2010$Time, times_2011$Time)
```


### Try it out

Build a 90% CI for the difference in mean time between female and male 50-54 year old runners in 2010.

```{r}
bm_activity <- bm %>% 
  filter(Year == 2010, Age_Range == "50-54") %>% 
  group_by(Sex) %>% 
  summarize(avgTime = mean(Time), sdTime = sd(Time), n = n())

bm_activity
```



## H-test for a difference in means

Is the change in the average time for an 18-34 year old female runner from 2010 to 2011 more than 0? Let's calculate a test statistic.

```{r}
bm_twomeans_summary <- bm %>% 
  filter(Sex == "female", Age_Range == "18-34") %>% 
  group_by(Year) %>% 
  summarize(avgTime = mean(Time), sdTime = sd(Time), n = n())

xbar1 <- bm_twomeans_summary$avgTime[1]
xbar2 <- bm_twomeans_summary$avgTime[2]
s1 <- bm_twomeans_summary$sdTime[1]
s2 <- bm_twomeans_summary$sdTime[2]
n1 <- bm_twomeans_summary$n[1]
n2 <- bm_twomeans_summary$n[2]

se <- sqrt(s1^2/n1 + s2^2/n2)

test_stat <- ((xbar1 - xbar2) - 0)/se

test_stat
```

Is the value 0.316 plausible under a T with w = 7134 degrees of freedom?

```{r}
w_numer <- (s1^2/n1 + s2^2/n2)^2
w_denom <- (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))

w <- w_numer / w_denom

gt(df = w) +
  geom_vline(xintercept = test_stat, color = "red", lty = 2)
```

For our one-sided hypotheses, the p-value is the area below our observed test statistic.

```{r}
gt(df = w) +
  geom_vline(xintercept = test_stat, color = "red", lty = 2) +
  geom_t_fill(df = w, b = test_stat, fill = "dodgerblue")
```

```{r}
pt(0.316, df = 2)
```

Let's also recreate the result with `t.test`.  We need to specify a direction.

```{r}
times_2010 <- bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2010)

times_2011 <- bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2011)

t.test(times_2010$Time, times_2011$Time, alternative = "less")
```


## Paired T test

We want to test whether the difference (growth) in height from age 13 to 14 is positive.

```{r}
thirteen <- c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen <- c(46.3, 60.5, 68.2, 59.4, 50.6)

# This works the way we want it to
growth <- fourteen - thirteen
```

```{r}
n <- length(growth)
# Note: n is quite small here, we only have 5 individuals; so using the t-distribution instead of the normal is important!

xbar <- mean(growth)
s <- sd(growth)

test_stat <- (xbar - 0)/(s/sqrt(n))

test_stat
```

We have an observed test statistic of 5.39, and a null T with 4 df.

```{r}
gt(df = n -1) +
  geom_vline(xintercept = test_stat, color = "red", lty = 2)
```

The p-value is the area above our test statistic.

```{r}
pt(test_stat, df = n-1, lower.tail = F)
```


### Check in R

```{r}
t.test(growth, alternative = "greater")

t.test(fourteen, thirteen, paired = T, alternative = "greater")
```

If we don't pair tthe data, we get a non-significant result.

```{r}
t.test(fourteen, thirteen, alternative = "greater")
```







