---
title: "Bob-Ross"
output: html_document
date: "2024-11-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("tidyverse")
```

# Sec 006 Monday 11/25

```{r}
all_elements <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/elements-by-episode.csv")
```

```{r}
head(all_elements)
```
```{r}
all_elements %>% count(TITLE) %>% filter(n>1)
```
```{r}
painting_elements <- all_elements %>% filter(GUEST==0) %>% select(-c(DIANE_ANDRE, STEVE_ROSS, GUEST, contains("FRAME")))

painting_elements
```
```{r}
painting_elements %>% 
  summarize(across(AURORA_BOREALIS:WINTER, mean))
```
```{r}
all_elements %>% summarise(across(contains("FRAME"), mean))
```
```{r}
graph <- painting_elements %>% 
  pivot_longer(-c(EPISODE, TITLE), 
               names_to = "feature", values_to = "appeared") %>% group_by(feature) %>% summarise(mean = round(mean(appeared)*100)) %>% arrange(desc(mean))

graph                      
```

```{r}
graph %>% filter(mean > 2) %>%  ggplot(mapping = aes(x = mean, y = feature)) + geom_col(  fill = "skyblue")

graph %>% filter(mean > 2) %>% ggplot(mapping = aes(x = mean, y = reorder(feature, mean))) + geom_col(  fill = "skyblue")

graph %>% filter(mean >= 2) %>% ggplot(mapping = aes(x = mean, y = reorder(feature, mean))) + geom_col(  fill = "skyblue") + geom_text(aes(label = mean), hjust = "left")
```
```{r}
graph %>% filter(mean > 0.02) %>% ggplot(mapping = aes(x = mean, y = reorder(feature, mean))) + 
  geom_col(  fill = "dodgerblue") +
  geom_text(aes(label = round(mean*100)), hjust = "left") +
  labs(
    title = "The Paintings of Bob Ross",
    subtitle = "Percentage containing each element",
    x = "",
    y = ""
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    plot.background = element_rect(fill = "gray96", color = "gray96"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  )

```


# Sec 003 Monday 11/25

```{r}
library("tidyverse")
```


```{r}
all_elements <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/elements-by-episode.csv")
```

```{r}
all_elements %>% count(TITLE) %>% filter(n>1) 
all_elements %>% filter(TITLE == "\"LAKESIDE CABIN\"" | TITLE == "\"MOUNTAIN WATERFALL\"")
```

```{r}
painting_elements <- all_elements %>% filter(GUEST==0) %>%  select(- c(DIANE_ANDRE, STEVE_ROSS, GUEST, contains("FRAME")))

painting_elements
```

```{r}
painting_elements %>% summarise(across(AURORA_BOREALIS:WINTER,mean))
```

```{r}
graph <- painting_elements %>% 
  pivot_longer(-c(EPISODE, TITLE), 
               names_to = "feature", values_to = "appeared") %>% group_by(feature) %>% summarise(mean = round(mean(appeared)*100)) %>% arrange(desc(mean))
graph
```


```{r}
graph %>% filter(mean>=2) %>% ggplot(mapping = aes(y = feature, x = mean)) + geom_col()

graph %>% filter(mean>=2) %>% ggplot(mapping = aes(y = reorder(feature, mean), x = mean)) + 
  geom_col(fill = "dodgerblue") + geom_text(aes(label = mean), hjust = "left") +
  labs(title = "The Paintings of Bob Ross", subtitle = "Percentage containing each element", x = "",
    y = "") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    plot.background = element_rect(fill = "gray96", color = "gray96"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  )
```


# Sec 005 Monday 11/25


```{r}
library("tidyverse")

all_elements <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/elements-by-episode.csv")
```
```{r}
all_elements %>% count(TITLE) %>% filter(n>1)

all_elements %>% filter(TITLE == "\"LAKESIDE CABIN\"" | TITLE == "\"MOUNTAIN WATERFALL\"") %>% select(EPISODE, TITLE)
```

```{r}
painting_elements <- all_elements %>% filter(GUEST == 0) %>% select(-c(DIANE_ANDRE, STEVE_ROSS, GUEST, contains("FRAME")))

painting_elements
```

```{r}
painting_elements %>% summarise(across(AURORA_BOREALIS:WINTER, mean))
```

```{r}
graph <- painting_elements %>% 
  pivot_longer(-c(EPISODE, TITLE), 
               names_to = "feature", values_to = "appeared") %>% group_by(feature) %>% summarise(mean = round(mean(appeared)*100)) %>% arrange(desc(mean))
graph
```

```{r}
graph %>% filter(mean >= 2) %>% ggplot(mapping = aes(y = feature, x = mean)) + geom_col()

graph %>% filter(mean >= 2) %>% ggplot(mapping = aes(y = reorder(feature, mean), x = mean)) + geom_col(fill = "dodgerblue") + geom_text(aes(label = mean)) + labs(title = "The Paintings of Bob Ross", subtitle = "Percentage containing each element", x = "",
    y = "") +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_blank(),
    plot.background = element_rect(fill = "gray96", color = "gray96"),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  )
```

# Sec 006 Monday 12/02

```{r}
painting_colors <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/bob_ross_paintings.csv")
```
```{r}
ggplot(data = painting_colors) + geom_bar(aes(x = num_colors))
```


```{r}
painting_colors %>% 
  filter(num_colors == 1) %>% 
  relocate(img_src, youtube_src, season, episode, painting_title)
```

```{r}
painting_colors %>%
  mutate(seasonChar = ifelse(season < 10, str_c("S0", season), str_c("S", season)),
         episodeChar = ifelse(episode < 10, str_c("E0", episode), str_c("E", episode)),
         EPISODE = str_c(seasonChar, episodeChar)) %>% 
  select(season, episode, seasonChar, episodeChar, EPISODE)


joined <- painting_colors %>%
  mutate(seasonChar = ifelse(season < 10, str_c("S0", season), str_c("S", season)),
         episodeChar = ifelse(episode < 10, str_c("E0", episode), str_c("E", episode)),
         EPISODE = str_c(seasonChar, episodeChar)) %>% 
  full_join(all_elements, join_by(EPISODE)) %>% select(ID = EPISODE, painting_title, num_colors, AURORA_BOREALIS:WINTER, Black_Gesso:Alizarin_Crimson)
```


## Single proportion for man-made structures

```{r}
structure_summary = joined %>% 
  summarize(structures = sum(STRUCTURE),
    n = n())

structure_summary
```

```{r}
p_hat_AC <- (85+2)/(403+4)

cv <- qnorm(0.975)
se <- sqrt((p_hat_AC*(1-p_hat_AC))/(403+4))

p_hat_AC + c(-1,1)*cv*se
```

```{r}
p_hat <- 85/403
cv <- qnorm(0.975)
se <- sqrt((p_hat*(1-p_hat))/(403))

p_hat + c(-1,1)*cv*se
```

# Sec 003 Monday 12/02

```{r}
painting_colors <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/bob_ross_paintings.csv")
```

```{r}
ggplot(data = painting_colors) + geom_bar(aes(x = num_colors)) + scale_x_continuous(breaks = 0:15)
```


```{r}
painting_colors %>% filter(num_colors == 1)
```

```{r}

```

```{r}
joined <- painting_colors %>%
  mutate(seasonChar = ifelse(season < 10, str_c("S0", season), str_c("S", season)),
         episodeChar = ifelse(episode < 10, str_c("E0", episode), str_c("E", episode)),
         EPISODE = str_c(seasonChar, episodeChar)) %>% 
  full_join(all_elements, join_by(EPISODE))

head(joined)
```

## Single proportion:

What is the true proportion of man made structures in the paintings?

```{r}
structure_summary = joined %>% 
  summarize(structures = sum(STRUCTURE),
    n = n())

structure_summary
```
95% Wald confidence interval

```{r}
p_hat <- 85/403
cv <- qnorm(0.975)
se <- sqrt(p_hat*(1 - p_hat)/403)

p_hat + c(-1,1)*cv*se
```

95% Agrestti-Coull CI

```{r}
p_hat_AC <- (85+2)/(403+4)
cv <- qnorm(0.975)
se <- sqrt((p_hat_AC*(1-p_hat_AC))/(403+4))

p_hat_AC + c(-1,1)*cv*se
```

# Sec 005 Monday 12/2

```{r}
painting_colors <- read_csv("~/Desktop/Stat240/FA24_STAT 240/data/bob_ross_paintings.csv")
```

```{r}
painting_colors %>% ggplot(aes(x = num_colors)) + geom_bar() + scale_x_continuous(breaks = 0:15)
```


```{r}
painting_colors %>% filter(num_colors == 1)
```

```{r}
joined <- painting_colors %>% mutate(seasonChar = ifelse(season<10, str_c("S0", season), str_c("S",season)), episodeChar = ifelse(episode<10, str_c("E0", episode), str_c("E",episode)), EPISODE = str_c(seasonChar, episodeChar)) %>% full_join(all_elements, join_by(EPISODE))
```


## Single proportion

What is the true proportion of TJOP episodes featuring a man-made structure

```{r}
joined %>% summarise(x = sum(STRUCTURE), n = n())
```

$X = \text{number of times the TJOP episode has a man-made structure} \\ X \sim Binom(403, p)$

95% Confidence Interval:

```{r}
cv <- qnorm(0.975)
```


Wald Adjustment

```{r}
p_hat <- 85/403
se <- sqrt(p_hat*(1-p_hat)/403)

p_hat + c(-1,1)*cv*se
```

Agrestti-Coull Adjustment:

```{r}
p_hat_AC <- (85+2)/(403+4)
cv <- qnorm(0.975)
se <- sqrt((p_hat_AC*(1-p_hat_AC))/(403+4))

p_hat_AC + c(-1,1)*cv*se
```


# Sec 006 Wednesday 12/4

- Hypothesis testing

- Step 2: State the hypothesis:

$$H_0: p=0.2, H_{\alpha}: p \neq 0.2$$
- Step 3: Test statistic and null distribution

$$X_{H_0} \sim Binom(403, 0.2)$$
- Step 4: Calculate the p-value

```{r}
p_hat <- 85/403
p_hat

t <- (p_hat - 0.2)/sqrt(0.2*0.8/403)
t
2*(1-pnorm(t)) # upper tail p-value


x <- 85
1-pbinom(85-1, 403, 0.2) + pbinom(75, 403, 0.2)
```
- Step 6: Given that $\alpha = 0.05$, what conclusion can you make?


## Difference of proportion

Is there a significant difference in the proportion of times Bob paints a man-made structure compared to a guest painting a man-made structure

```{r}
joined %>% group_by(GUEST) %>% summarise(n = n(), x = sum(STRUCTURE))
```


- 95% Agrestti-Coull Confidence Interval

```{r}
x1 <- 84
n1 <- 381
x2 <- 1
n2 <- 22

p1_hat_AC <- (x1+1)/(n1+2)
p2_hat_AC <- (x2+1)/(n2+2)

(p1_hat_AC-p2_hat_AC) + c(-1,1)*(qnorm(0.975))*(sqrt(p1_hat_AC*(1-p1_hat_AC)/(n1+2)+p2_hat_AC*(1-p2_hat_AC)/(n2+2)))
```


```{r}
p_value <- 2*(1-pnorm(1.956527)) # 2*(pnorm(-1.956527))
p_value
```


# Sec 006

- Step 2: State the hypothesis

$$H_0: p= 0.2, H_{\alpha}: p\neq 0.2 $$

- Step 3: Test statistic and null distrubition

$$X_{H_0} \sim Binom(403,0.2)$$

```{r}
p_hat <- 85/403

z <- (p_hat - 0.2)/sqrt(0.2*0.8/403)
z

2*(1-pnorm(z)) # p value using the test statistic 

x <- 85
pbinom(85-1, 403, 0.2, lower.tail = F) + pbinom(75, 403, 0.2) # p value using the null distribution of X
```


## Difference of proportion

The propotion of painting with a man-made structure when we had a guest or no guest.


```{r}
joined %>% group_by(GUEST) %>% summarise(n = n(), x = sum(STRUCTURE))
```
- Build a 95% Agrestti-Coull Confidence Interval

```{r}
p1_hat_AC <- (84+1)/(381+2)
p2_hat_AC <- (1+1)/(22+2)

se <- sqrt(p1_hat_AC*(1-p1_hat_AC)/(n1+2) + p2_hat_AC*(1-p2_hat_AC)/(n2+2))

(p1_hat_AC-p2_hat_AC) + c(-1,1)*qnorm(0.975)*se
```


- Hypothesis Testing

```{r}
2*(1-pnorm(1.956527))
```

# Sec 005 Wednesday 12/4

- Step 2: State the hypothesis:

$$H_0: p= 0.2, H_{\alpha}: p\neq 0.2 $$
- Step 3: Test Statistic and the null distribution

```{r}
z <- (85/403 - 0.2)/sqrt(0.2*0.8/403)
z

2*(1- pnorm(z))

x <- 85
pbinom(85-1, 403, 0.2, lower.tail = F) + pbinom(75, 403, 0.2) # p value using the distribution of X
```

$\alpha = 0.05$

Conclusion: p=0.2


## Difference of proportion

Is there difference in the proportion of paintings with man-made structure when Bob is painiting vs a guest is painting

```{r}
joined %>% group_by(GUEST) %>% summarise(x = sum(STRUCTURE), n = n())
```

```{r}
p1_hat <- 84/381
p2_hat <- 1/22
```

- 95% Confidence interval using the Agrestti-Coull adjustment.

- Hypothesis testing:

```{r}
z <- 1.956527

2*(1-pnorm(z))

2*pnorm(-z)
```


```{r}
t.test(painting_colors$num_colors, mu = 10.5)
t.test(painting_colors$num_colors, mu = 10.5, alternative = "greater")
```

- Hypothesis testing for difference of means

```{r}
joined %>% group_by(GUEST) %>% summarize(xbar = mean(num_colors), sd = sd(num_colors), n = n())
```

```{r}
guest = joined %>% filter(GUEST==1)
bob = joined %>%filter(GUEST==0)

t.test(bob$num_colors, guest$num_colors, conf.level = 0.90)

t.test(num_colors~GUEST, joined)
```

```{r}
total_features_vector = joined %>%
  select(AURORA_BOREALIS:WINTER) %>%
  select(!contains("FRAME")) %>%
  rowSums()

paintings_joined = joined %>%
  mutate(num_features = total_features_vector)

model <- lm(num_features ~ num_colors, paintings_joined)

summary(model)
```

```{r}
residual <- resid(model)

x <- paintings_joined$num_colors

ggplot(mapping = aes(x = x, y = residual)) + geom_point() + geom_smooth()
```

$$H_0: \beta_1 = 0, H_{\alpha}: beta_1 >0$$

```{r}
beta_hat_1 <- -0.04366
se = 0.04748
n = nrow(paintings_joined)
test_stat = (beta_hat_1 - 0)/se
test_stat

1 - pt(test_stat, n-2)
```


# Sec 003 Friday 12/6

Hypothesis testing 

$$H_0: \mu = 10.5, H_{\alpha}: \mu \neq 10.5$$

```{r}
t.test(joined$num_colors, mu = 10.5)
t.test(joined$num_colors, mu = 10.5, alternative = "greater")
```

- Difference of means

```{r}
joined %>% group_by(GUEST) %>% summarize(n = n(), xbar = mean(num_colors), sd = sd(num_colors))
```

- Hypothesis testing:

$$H_0: \mu_{Bob} = \mu_{guest}, H_{\alpha}: \mu_{Bob}>\mu_{guest}$$
```{r}
bob = joined %>% filter(GUEST==0)
guest = joined %>% filter(GUEST==1)
t.test(bob$num_colors, guest$num_colors, alternative = "greater")
```
```{r}
total_features_vector = joined %>%
  select(AURORA_BOREALIS:WINTER) %>%
  select(!contains("FRAME")) %>%
  rowSums()

paintings_joined = joined %>%
  mutate(num_features = total_features_vector)

model = lm(paintings_joined$num_features~paintings_joined$num_colors)
```


```{r}
summary(model)
```

```{r}
se = 0.04748
n = nrow(paintings_joined)
test_stat = (-0.04366 - 0)/se

pt(test_stat, df = n - 2, lower.tail = FALSE) # greater than 
pt(test_stat, df = n - 2) # less than
2*pt(test_stat, df = n - 2) # two sided
```

# Sec 005 Friday 12/6

Hypothesis testing

$$H_0: \mu = 10.5, H_{\alpha}: \mu \neq 10.5$$

```{r}
t.test(joined$num_colors, mu = 10.5)
```
```{r}
joined %>% group_by(GUEST) %>% summarise(n = n(), xbar = mean(num_colors), sd = sd(num_colors))
```

- Difference of means: 

$$H_0: \mu_{Bob} = \mu_{guest}, H_{\alpha}: \mu_{Bob} > \mu_{guest}} $$


```{r}
bob = joined %>% filter(GUEST==0)
guest = joined %>% filter(GUEST==1)

t.test(bob$num_colors, guest$num_colors)
t.test(num_colors~GUEST, joined)
```

```{r}
total_features_vector = joined %>%
  select(AURORA_BOREALIS:WINTER) %>%
  select(!contains("FRAME")) %>%
  rowSums()

paintings_joined = joined %>%
  mutate(num_features = total_features_vector)

 model = lm(paintings_joined$num_features~paintings_joined$num_colors)
```


```{r}
summary(model)
```

$$H_{alpha}: \beta_1>0$$

```{r}
t = -0.04366/0.04748
t

1 - pt(t, 401)
pt(t, 401)
```

















