---
title: "Discussion 12"
output: html_document
---

# Preliminaries

-   This file should be in `STAT240/discussion/ds12` on your local computer.

-   This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

-   This assignment is to be completed in the first 10-20 mins of the discussion session. The rest of the discussion time the students are to work on the group projects with their discussion groups.

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Members:

> Absent Group Members:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
```

## Problems

The questions in this discussion assignment use the boston marathon data set. The data set `bm_sample` contains a random sample of 100 men and 100 women, between the age of 35-39, who ran in and completed the 2010 Boston Marathon. Times are listed in minutes.

```{r}
# This code is provided for you
bm_sample = read_csv("../../data/bm_sample.csv")
```

### Problem 1

Create and display a table with a numerical summary of the data.
Calculate the sample size, mean time, standard deviation of the times, and standard error of the mean ($s/\sqrt{n}$) separately for each sex. This table will have one row for each sex.

```{r}
# Write your code here!
```

### Problem 2

Construct 95% confidence intervals for each sex seperately for the mean time to complete the 2010 Boston Marathon among all competitors of that sex aged 35-39.


```{r}
# Write your code here!
```


### Problem 3

**Part (a)** Perform hypothesis test at 10% significance level for the difference of mean or the two - sample `t.test()` for the mean time to complete the 2010 Boston Marathon for male is less than that for female participants. Write down all the steps of the hypothesis test.

```{r}
# Write your code here!
```


**Part (b)** Can you support the conclusion made for the hypothesis test based on the confidence interval obtained in the Step 4 and 5?

> Write your answer here!





