---
title: "Discussion 12"
output: html_document
---

# Preliminaries

-   This file should be in `STAT240/discussion/ds12` on your local computer.

-   This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

-   This assignment is to be completed in the first 10-20 mins of the discussion session. The rest of the discussion time the students are to work on the group projects with their discussion groups.

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Members:

> Absent Group Members:

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      error = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
library(tidyverse)
source("~/Desktop/STAT 240/scripts/ggprob.R")
```

## Problems

The questions in this discussion assignment use the boston marathon data set. The data set `bm_sample` contains a random sample of 100 men and 100 women, between the age of 35-39, who ran in and completed the 2010 Boston Marathon. Times are listed in minutes.

```{r}
# This code is provided for you
bm_sample = read_csv("~/Desktop/STAT 240/data/bm_sample.csv")
```

### Problem 1

Create and display a table with a numerical summary of the data.
Calculate the sample size, mean time, standard deviation of the times, and standard error of the mean ($s/\sqrt{n}$) separately for each sex. This table will have one row for each sex.

```{r}
bm_sum = bm_sample %>% 
  group_by(Sex) %>% 
  summarize(n = n(), mean = mean(Time), sd = sd(Time)) %>% 
  mutate(se = sd/sqrt(n))

bm_sum
```

### Problem 2

Construct 95% confidence intervals for each sex seperately for the mean time to complete the 2010 Boston Marathon among all competitors of that sex aged 35-39.


```{r}
# 95% CI for Male participants
mean_male <- bm_sum$mean[2]
se_male <- bm_sum$se[2]
cv <- qt(0.975, 99)

mean_male + c(-1,1)*cv*se_male


# Other way
men = bm_sample %>% 
  filter(Sex == "male") %>% 
  pull(Time)

t.test(men)
```


```{r}
# 95% CI for Female participants

mean_female <- bm_sum$mean[1]
se_female <- bm_sum$se[1]
cv <- qt(0.975, 99)

mean_female + c(-1,1)*cv*se_female

# Other way
women = bm_sample %>% 
  filter(Sex == "female") %>% 
  pull(Time)

t.test(women)
```

### Problem 3

**Part (a)** Perform hypothesis test at 10% significance level for the difference of mean or the two - sample `t.test()` for the mean time to complete the 2010 Boston Marathon for male is less than that for female participants. Write down all the steps of the hypothesis test.

> Solution:

- Step 1: Model Statement: Let $X$ be the time taken by a male participant to complete the 2010 Boston Marathon and $Y$ be the time taken by a female participant to complete the 2010 Boston Marathon

$$X \sim D_{x}(\mu_{x},\sigma_{y}) \\
Y \sim D_{y}(\mu_{y},\sigma_{y})$$

Assumption: $X$ and $Y$ are independent of each other.

- Step 2: State hypothesis:

$$
H_0: \mu_{x} = \mu_{y} \implies \mu_{x} - \mu_{y} = 0 \\
H_{\alpha}: \mu_{x} < \mu_{y} \implies \mu_{x} - \mu_{y} < 0
$$

- Step 3: Test Statistics and null distribution.

$$t = \frac{\bar{x} - \bar{y}}{\sqrt{\frac{s_x^2}{n_x} + \frac{s_y^2}{n_y}}} \sim t(w)$$

where, $w$ is the Welch degree of freedom.

- Step 4 and 5: Get the test statistic and the p-value for the dataset using function `t.test()`

```{r}
t.test(men, women, alternative = "less", conf.level = 0.90)
```

- Step 6: Make conclusion for $\alpha = 0.10$

We reject $H_0$ or we have enough evidence to support that the mean time taken by male participants to finish the 2010 Boston Marathon is less than that of female participants.

**Part (b)** Can you support the conclusion made for the hypothesis test based on the confidence interval obtained in the Step 4 and 5?

> Yes, since the 95% confidence interval doesn't contain 0 and covers values less that 0 suggesting that the difference of the means is going to be less than 0.





