---
title: "Section 9: Binomial Distribution"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```

## Binomial properties

Use `choose` or `factorial` to count the number of ways to have x successes in n total trials.

```{r}
n <- 3
x <- 2

choose(n, x)

factorial(n) / (factorial(x) * factorial(n - x))
```


We can use `dbinom` to calculate a specific binomial probability.

```{r}
dbinom(2, 3, 0.6)
```

Here, it is used to fully describe a binomial distribution.

```{r}
n <- 3
p <- 0.6

tibble(
  x = 0:n,
  p = dbinom(x, n, p)
)
```

The function `pbinom` calculates a cumulative probability.  What is the probability that we hit 2 or fewer green lights?

```{r}
dbinom(0, n, p) + dbinom(1, n, p) + dbinom(2, n, p)

sum(dbinom(0:2, n, p))

pbinom(2, n, p)
```

### pbinom exercise

Let Y ~ Binom(8, 0.3).  Use `pbinom` to answer the following:

```{r}
# P(Y > 4)


# P(Y >=)


# P(3 <= Y <= 6)


```


## Mean and variance

The binomial shortcuts for mean and variance give us the same values as the standard formulas.

```{r}
lights <- 0:3
light_probs <- c(0.064, 0.288, 0.432, 0.216)

lights_mean <- sum(lights * light_probs)

lights_var <- sum((lights - lights_mean)^2 * light_probs)
```

We can also verify this with simulation.

```{r}
# rbinom generates a random binomial sample

rbinom(3, n, p)

large_sample <- rbinom(500000, n, p)

mean(large_sample)   # compare to 1.8
var(large_sample)    # compare to 0.72

tibble(x = large_sample) %>%  
  ggplot(aes(x)) +
  geom_histogram(stat = "count", color = "black",) +
  stat_count(geom = 'text',    # Add counts to each bin
             color = 'white', 
             aes(label = after_stat(count)),
           position = position_stack(vjust = 0.5)) +

  labs(
    title = "Simulated Binomial Distribution",
    subtitle = "n = 3, p = 0.6"
  )
```


## Quantiles

Consider an example where we flip a fair coin 10 times, and count the number of heads.

`pbinom` can be used to find a cumulative probability.

```{r}
n <- 10
p <- 0.5

# How much probability is at or below 5 on Binom(10, 0.5)?
pbinom(5, n, p)

# How much probability is at or below 6?
pbinom(6, n, p)
```

For `qbinom`, we input a cumulative probability and are given the corresponding x value.  This will return discrete x values.

```{r}
# To capture at least 70% of the probability in Binom(10, 0.5)
# How large do we need to make x?

qbinom(0.7, n, p)

# How "far in" do we need to go to capture at least 95% of the probability?
qbinom(0.95, n, p)

pbinom(7, n, p)
pbinom(8, n, p)
```

### Try it out!

Let X ~ Binom(90, 0.7)

```{r}
# Find the mean and sd of X


# Find P(X <= mu)


# Find P(mu - sigma <= X <= mu + sigma)


# Find the 5th and 95th percentiles of X


```





