---
title: "Section 8: Probability Basics"
author: "Student's Name Here"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```

## Random variables

For a RV that can take on a few specific values, we can visually represent the variable with line segments.  The y-axis corresponds to probability.

```{r}
apgar_vals <- 0:10
apgar_probs <- c(0.001, 0.006, 0.007, 0.008, 0.012, 0.02,
                 0.038, 0.099, 0.319, 0.437, 0.053)

apgar <- tibble(apgar_vals, apgar_probs)

apgar_distribution <- ggplot(apgar, aes(x = apgar_vals, y = apgar_probs)) +
  geom_segment(aes(xend = apgar_vals, yend = 0), color = "blue", size=2) +
  geom_hline(yintercept = 0) +
  ylab("P(X=x)") +
  ggtitle("Distribution of Apgar Scores")

apgar_distribution
```


## Properties of RVs

What is the mean (expected value), variance, and SD of the Apgar score data?

```{r}
apgar_mean <- sum(apgar_vals*apgar_probs)

apgar_var <- sum((apgar_vals - apgar_mean)^2 * apgar_probs)

apgar_sd <- sqrt(apgar_var)

apgar_mean; apgar_var; apgar_sd
```


Let X be a discrete RV with possible values (0, 1, 2, 3, 4).  A partial distribution is given below.

```{r}
x_dist <- tibble(
  x = 0:4,
  p = c(0.15, 0.25, 0.05, 0.35, NA)
)

x_dist
```

Find the missing value for P(X = 4) and update the table.  Try to do this with dplyr commands.

```{r}

```

Next, find E(X), V(X), and sd(X).

```{r}

```

Finally, use ggplot with `geom_segment` to visualize the probability distribution of X.

```{r}

```







