---
title: "Section 4: ggplot2 Basics"
author: "Student's Name Here"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```


## Mendota data

Use the code below to load in the Lake Mendota freeze data.  The .csv file must be in your `data` folder.

```{r}
## This assumes that:
### STAT240/data/ contains the data file
### STAT240/lecture/sect04-ggplot-basics/ is your working directory.
### If this gives you "Error: could not find file ... in working directory ...", go to Session > Set Working Directory > To Source File Location, and try again.
### If that doesn't work, then you downloaded one or both files to the wrong place, or they have the wrong name - make sure they don't have a " (1)" or "-1" at the end of their names, which can happen when you download multiple times.

mendota <- read_csv("../../data/lake-mendota-winters-2023.csv")
```

Now explore the data.

```{r}
glimpse(mendota)
```


## Grammar of graphics

```{r}
# Basic empty canvas
ggplot()

# We defined the mapping - where is the plot?
ggplot(mendota, aes(x = year1, y = duration))
```

To actually build the plot, we need to specify a geom and add it to the plot.

```{r}
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point()

ggplot(mendota, aes(x = year1, y = duration)) +
  geom_line()

ggplot(mendota, aes(x = year1, y = duration)) +
  geom_line() +
  geom_point()
```

## Customization

We can customize several aspects of our plot.

```{r}
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point(size = 5, color = "red")

ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point(shape = 8, alpha = 0.5)
```


Let's add a third variable into our mapping.

```{r}
ggplot(mendota, aes(x = year1, y = duration, col = intervals)) +
  geom_point()
```

We want to be treating this as a categorical variable, not a continuous one.  Categorical variables in R are called "factors".

```{r}
ggplot(mendota, aes(
    x = year1, 
    y = duration, 
    color = as.factor(intervals)
    ) # this parentheses ends the aes() call
  ) + # this parentheses ends the ggplot() call
  geom_point()


ggplot(mendota, aes(
    x = year1, 
    y = duration, 
    color = as.factor(intervals)
    ) 
  ) + 
  geom_point(size = 5)
```

Consider a basic ggplot with both a point and a line layer.

```{r}
ggplot(mendota, aes(x = year1, y = duration)) + 
  geom_point() +
  geom_line()
```

Change the aesthetics inside `geom_point` and `geom_line` to distinguish the two layers.

## Variable aesthetics

In the following plot, the aesthetic of `intervals` applies to both the lines and the dots.

```{r}
ggplot(mendota, aes(x = year1, y = duration, color = intervals)) +
  geom_line(size = 1.5) +
  geom_point(size = 2)
```

If we only want the dots to be colored, we have to specify the color mapping within `geom_point` rather than the original `ggplot`.

```{r}
ggplot(mendota, aes(x = year1, y = duration) ) + 
  geom_line(size = 1.5) + 
  geom_point(mapping = aes(color = intervals), size = 2) 
```

What's wrong with the following plots?  How can we fix them?

```{r}
# Read and interpret the error messages from these two plots
ggplot(mendota) +
  geom_point(aes(color = intervals))

ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point(color = intervals)
```

```{r}
# Why are these points not huge? Why is there a legend for "size"?
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point(aes(color = intervals, size = 1000))

# This just produces a gridded canvas with no points. Why?
ggplot(mendota, aes(x = year1, y = duration, color = intervals))
  geom_point()
```




