---
title: "Section 5: More ggplot2"
author: "Student's Name Here"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
```


## Mendota data

Use the code below to load in the 2024 Lake Mendota freeze data.  The .csv file must be in your `data` folder.

```{r}
## This assumes that:
### STAT240/data/ contains the data file
### STAT240/lecture/sect04-ggplot-basics/ is your working directory.
### If this gives you "Error: could not find file ... in working directory ...", go to Session > Set Working Directory > To Source File Location, and try again.
### If that doesn't work, then you downloaded one or both files to the wrong place, or they have the wrong name - make sure they don't have a " (1)" or "-1" at the end of their names, which can happen when you download multiple times.

mendota <- read_csv("../../data/lake-mendota-winters-2023.csv")
```

Now explore the data.

```{r}
glimpse(mendota)
```


## More geoms

`geom_smooth` gives us a good idea of the general trend in a scatterplot.

```{r}
ggplot(data = mendota, mapping = aes(x = year1, y = duration) ) +
  geom_point()

ggplot(data = mendota, mapping = aes(x = year1, y = duration) ) +
  geom_point() +
  geom_smooth()
```

The gray ribbon represents a confidence interval, and can be omitted.  We can also use different methods to obtain a line.

```{r}
ggplot(data = mendota, mapping = aes(x = year1, y = duration) ) +
  geom_point() +
  geom_smooth(se = FALSE)

ggplot(data = mendota, mapping = aes(x = year1, y = duration) ) +
  geom_point() +
  geom_smooth(se = FALSE, method = "lm")
```


Histograms, density plots, and box plots are useful for visualizing a single variable.

```{r}
ggplot(mendota, aes(x = duration)) + 
  geom_histogram()

# Can customize the look
ggplot(mendota, aes(x = duration)) + 
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  )
```

We can use `binwidth, bins, center`, and `boundary` to custoimze the bins.  These two plots have the same output:

```{r}
ggplot(mendota, aes(x = duration)) + 
  geom_histogram(
    binwidth = 5,
    boundary = 70,
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  
  # This is just labelling the x-axis to make it clearer where the bins are. 
  # We'll discuss this in more depth later.
  scale_x_continuous(breaks = seq(30, 160, by = 10))

ggplot(mendota, aes(x = duration)) + 
  geom_histogram(
    binwidth = 5,
    center = 72.5,
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  scale_x_continuous(breaks = seq(30, 160, by = 10))
```

Try to get a good balance - not too many and not too few bins!

```{r}
ggplot(mendota, aes(x = duration)) +
  geom_histogram(
    bins = 100,
    color = "steelblue4",
    fill = "skyblue1")

ggplot(mendota, aes(x = duration)) +
  geom_histogram(
    bins = 5,
    color = "steelblue4",
    fill = "skyblue1")
```


Density plots are a "smooth" version of a histogram.

```{r}
ggplot(mendota, aes(duration)) +
  geom_density(
    # Notice the same border-inside coloring scheme with color vs. fill
    color = "red4",
    fill = "red", # Default is totally clear, no fill color
    size = 2
  )

# Overlaying both plots
ggplot(mendota, aes(duration)) +
  geom_histogram(
    aes(y = after_stat(density)), # This line shrinks the histogram's height to be on the same scale as geom_density; out of scope of this course, no need to worry about it
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  geom_density(
    color = "red4",
    fill = "red1",
    alpha = 0.3, # Transparency... the density plot is "on top"
    size = 2
  )
```


Boxplots visualize the quartiles of our data.

```{r}
ggplot(mendota, aes(x = duration)) +
  geom_boxplot()
```

Let's make a categorical variable for century.

```{r}
# You'll learn more about this code next week!
mendota <- mendota %>% 
  mutate(century = as.character(floor(year1/100)+1),
         century = case_when(
           century == "19" ~ "19th",
           century == "20" ~ "20th",
           century == "21" ~ "21st"
         ))
```

Now let's re-create the one-variable plots, but color by century.

```{r}
ggplot(mendota, aes(x = duration)) +
  geom_histogram()

ggplot(mendota, aes(x= duration)) +
  geom_density()

ggplot(mendota, aes(x = duration)) +
  geom_boxplot()
```


Let's see how we can annotate plots with lines and text.

```{r}
meanDuration <- mean(mendota$duration)

ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  geom_vline(xintercept = meanDuration,
             size = 2) 

ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  geom_vline(xintercept = meanDuration,
             size = 1.5) +
  geom_text(aes(x = 78, y = 20,
                label = "Mean Freeze Duration"),
                stat = "unique")
```

We can also add multiple lines to a plot.

```{r}
useful_values <- meanDuration +
  c(-3, -2, -1, 1, 2, 3) * sd(mendota$duration)

useful_values

ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  geom_vline(xintercept = useful_values,
             size = 1.5) 
```


Here's an example of using lines to identify and visualize outliers.

```{r}
iqr <- IQR(mendota$duration)
first_q <- quantile(mendota$duration, 0.25)
third_q <- quantile(mendota$duration, 0.75)

ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point() +
  geom_hline(
    yintercept = c(first_q - 1.5*iqr, third_q + 1.5*iqr)
    )
```


## Bar graphs

Bar graphs depict a single categorical variable.  `geom_bar` counts the number of instances in a dataset.

```{r}
ggplot(mendota, aes(century)) +
  geom_bar(
    color = "steelblue4",
    fill = "skyblue1"
  )
```

With `geom_col`, you specify the heights of the bars yourself.  It is more flexible than `geom_bar` but more difficult to use.

Let's make a new version of the Lake Mendota data, that has only three rows (one for each century).

```{r}
# We'll learn more about this code next week!
mendota_summarized <- mendota %>% 
  group_by(century) %>% 
  summarize(numYears = n(), avgDuration = mean(duration))
  
mendota_summarized
```

Now, we can use `geom_col` with a y-aesthetic.

```{r}
# This code recreates the geom_bar graph
ggplot(mendota_summarized, aes(century, numYears)) +
  geom_col(
    color = "steelblue4",
    fill = "skyblue1"
  )

# But we can make other plots as well
ggplot(mendota_summarized, aes(century, avgDuration)) +
  geom_col(
    color = "steelblue4",
    fill = "skyblue1"
  )
```


## More customization

### Scales

We can define more useful scales that override the ggplot defaults.

```{r}
# Notice ggplot's default x-axis choices
ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  )


ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  scale_x_continuous(
    breaks = c(30, 90, 150),
    labels = c("1 month", "3 months", "5 months"),
    limits = c(15, 165),
    minor_breaks = NULL, # This specifies not to draw any vertical axis lines between the labeled points
  )
```

```{r}
ggplot(mendota, aes(duration)) +
  geom_histogram(
    color = "steelblue4",
    fill = "skyblue1"
  ) +
  scale_x_continuous(
    breaks = c(30, 90, 150),
    labels = c("1 month", "3 months", "5 months"),
    limits = c(-100, 300),
    minor_breaks = NULL
  ) +
  # Can you figure out what this addition is doing to the y-axis?
  scale_y_continuous(
    expand = expansion(mult = c(0,0.1)),
    limits = c(-10, 100)
  )
```

### Color Schemes

The default color scheme can be difficult to distinguish for certain types of color blindness.

```{r}
ggplot(mendota, aes(x= duration, fill = century)) +
  geom_density(alpha = 0.3)
```

We can use `viridis` palettes for either the fill or the color aesthetic.

```{r}
ggplot(mendota, aes(x= duration, fill = century)) +
  geom_density(alpha = 0.3) +
  scale_fill_viridis_d()

ggplot(mendota, aes(x= duration, fill = century)) +
  geom_density(alpha = 0.3) +
  scale_fill_viridis_d(option = "inferno")
```

We can also make our own custom color scheme.

```{r}
ggplot(mendota, aes(x= duration, fill = century)) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(
    values = c("19th" = "dodgerblue",
               "20th" = "peachpuff",
               "21st" = "mediumorchid")
    )
```


### Labels

The `labs` function can be used to label any part of the plot.

```{r}
density_plot <- ggplot(mendota, aes(x= duration, fill = century)) +
  labs(
    title = "Distribution of Freeze Duration by Century",
    subtitle = "Lake Mendota, 1855-2023",
    caption = "STAT 240",
    
    x = "Duration (in days)",
    y = "Density",
    fill = "Century"
  ) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(
    values = c("19th" = "dodgerblue",
               "20th" = "peachpuff", 
              "21st" = "mediumorchid")
    )

density_plot
```

```{r}
# What is the difference between these two plots?

ggplot(mendota, aes(x= duration, fill = century)) +
  labs(
    title = "Distribution of Freeze Duration by Century",
    subtitle = " ",
    caption = "STAT 240",
    
    x = "Duration (in days)",
    y = "Density",
    fill = "Century"
  ) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(
    values = c("19th" = "dodgerblue",
               "20th" = "peachpuff", 
              "21st" = "mediumorchid")
    )

ggplot(mendota, aes(x= duration, fill = century)) +
  labs(
    title = "Distribution of Freeze Duration by Century",
    subtitle = NULL,
    caption = "STAT 240",
    
    x = "Duration (in days)",
    y = "Density",
    fill = "Century"
  ) +
  geom_density(alpha = 0.3) +
  scale_fill_manual(
    values = c("19th" = "dodgerblue",
               "20th" = "peachpuff", 
              "21st" = "mediumorchid")
    )
```


### Themes

Themes change the background and overall look.

```{r}
density_plot +
  theme_minimal()

density_plot +
  theme_classic()

density_plot +
  theme(text = element_text(family = "Times New Roman", size = 16))
```

Let's try recreating a plot.

```{r}
teams <- c("Wolves", "Aston Villa", "Liverpool", "Tottenham", "Man City")
ESPN <- tibble(team = factor(teams, levels = teams),
       points = c(2.0, 2.0, 2.2, 2.4, 2.5))

ESPN

ggplot(ESPN)
```



## Faceting

Sometimes, it is easier to view variables as separate graphs rather than as aesthetics.

```{r}
ggplot(mendota, aes(x= duration, fill = century)) +
  geom_density(alpha = 0.3)

ggplot(mendota, aes(x = duration)) +
  geom_density() +
  facet_wrap(facets = vars(century))
```

`facet_grid` lets us facet based on two variables.

```{r}
ggplot(mendota, aes(x = duration)) +
  geom_density() +
  facet_grid(rows = vars(decade), cols = vars(century))

ggplot(mendota, aes(x = duration, fill = century)) +
  geom_density() +
  facet_grid(rows = vars(decade))
```

Let's make a column for whether a year is a leap year.

```{r}
mendota <- mendota %>% 
  mutate(leap_year = year1 %% 4 == 0)

mendota %>%
  select(year1, leap_year) %>% 
  head()
```

Leap years occur every century, so we can facet by both century and leap year.

```{r}
ggplot(mendota, aes(x = duration)) +
  geom_density() +
  facet_grid(rows = vars(century), cols = vars(leap_year))

# This plot is much harder to read:
ggplot(mendota, aes(x = duration, fill = century, linetype = leap_year)) +
  geom_density(alpha = 0.5, size = 1)
```


## Try it yourself!

The `economics` dataset automatically comes with the `ggplot` package.

```{r}
glimpse(economics)

economics <- economics %>%
  mutate(unemp_rate = unemploy / pop)

ggplot(economics, aes(x = date, y = unemp_rate)) +
  geom_line()
```

Make a graph of your own, and post it to the discussion thread in Canvas!












