---
title: "Homework 9"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE, error = TRUE,
                      fig.height = 3)
library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/homework/hw09` on your local computer.
- Download `happiness_2019.csv` and `dc_weather.csv` to `STAT240/data`.


# Problem 1

A one-sided hypothesis test of 

$$H_A: \beta_1 > 0$$

is performed on a linear model.  The p-value of this test is 0.035.  What would be the p-value if the same data was used to test the following alternatives?

**(a)** $$H_A: \beta_1 < 0$$

> Replace this text with your response.

**(b)** $$H_A: \beta_1 \neq 0$$

> Replace this text with your response.


# Problem 2

Continue working with the happiness index vs GDP of countries model from Homework 8.

```{r}
happiness <- read_csv("../../data/happiness_2019.csv")

# Remove spaces from column names
names(happiness) <- make.names(names(happiness))

happiness_mod <- lm(Score ~ GDP.per.capita, data = happiness)
```

Build and interpret a 98% CI for the true slope of the linear relationship between happiness index and GDP.  Does the interval cover 0?

```{r}
# Write your code here!
```


# Problem 3

Perform a hypothesis test of hypotheses

$$H_0: \beta_1 = 0 \quad \text{versus}\quad H_A: \beta_1 \neq 0$$

for the slope of the happiness model.  What is the test statistic, p-value, and conclusion at the 2% level?

```{r}
# Write your code here!
```


# Problem 4

How do the results of the hypothesis test in problem 3 relate to the results of the confidence interval in problem 2?

> Replace this text with your response.


# Problem 5

Which of the following conditions lead to a smaller standard error?  Briefly explain your choices.

- A smaller sample size vs a larger sample size

> Replace this text with your response.

- A smaller value of $\sigma$ vs a larger value of $\sigma$

> Replace this text with your response.

- Predicting closer to $\bar{x}$ vs predicting further from $\bar{x}$

> Replace this text with your response.

- A smaller variance vs a larger variance in the original $X$ data

> Replace this text with your response.


# Problem 6

Consider predicting the happiness index of a country with 1 GDP per capita.

**(a)** Build a *prediction interval* for the happiness index of a new country with $x^* = 1$ GDP per capita.

```{r}
# Write your code here!
```


**(b)** Build a *confidence interval* for the height of the regression line at $x^* = 1$.

```{r}
# Write your code here!
```

**(c)** Explain the difference in how the prediction and confidence intervals are calculated.

> Replace this text with your response.


# Problem 7

The file `dc_weather.csv` contains weather data for Washington, DC from August 2018 to August 2024. 

```{r}
weather <- read_csv("../../data/dc_weather.csv")

weather_mod <- lm(dew ~ tempmin, data = weather)
```

**(a)** Make a plot of minimum temperature (in C) on the x axis versus dew point on the y axis.  Comment on the shape of the data.

```{r}
# Write your code here!
```

> Replace this text with your response.

**(b)** Use R's `lm` to fit a linear model for dew point in terms of minimum temperature.  Perform a residual plot analysis to assess the validity of this model.

```{r}
# Write your code here!
```

> Replace this text with your response.


# Problem 8

We want to test whether the slope of the linear relationship between minimum temperature and dew point is **greater than 1**.  Write hypotheses corresponding to the question of interest and carry out the test on the weather data.  Make a conclusion with $\alpha = 0.05$.

```{r}
# Write your code here!
```

> Replace this text with your response.





