---
title: "Discussion 5"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
```

# Preliminaries

- This file should be in `STAT240/discussion/ds05`.
- You should also download `billboard_2000_weekly.csv` and `billboard_2000_dates.csv` to `STAT240/data`.

- This assignment should be completed in your **assigned Canvas groups**. Only one person has to submit on behalf of the group. You may submit whatever you have at the end of discussion section for full credit.

# Present Group Members

Please **write the names of the present group members** here.

> Present Group Member 1

> Present Group Member 2

> Present Group Member 3

> Present Group Member 4 (some groups only have 3)

# Background

```{r}
billboard_weekly = read_csv("../../data/billboard_2000_weekly.csv")
```

`billboard_weekly` contains data for 317 tracks which appeared in the Billboard Top 100 (most popular songs in the United States) at any time in the year 2000.

After the song enters the Billboard Top 100, its `rank` in the top 100 is recorded (1 is the most popular song of all) for the next 76 `week`s. 

If a song is still in the Top 100 in that `week` since its debut in the Top 100, its `rank` is recorded (1 is the most popular song of all).

Many do not last in the top 100 for long; `NA` means that song was not in the Billboard Top 100 in that week of its lifespan. 

For example, in the first ten rows of `billboard_weekly`: 2 Pac's "Baby Don't Cry" entered the Top 100 at #87 in February 2000, and peaked at #72 three weeks later. Eight weeks after it entered, it dropped out of the Top 100 entirely.

```{r}
billboard_weekly
```

# Problem 1

Using a `pivot` command, **create and save** a new dataframe called `billboard_tracks`, starting from `billboard_weekly`.

`billboard_tracks` should have one row PER TRACK (for 317 tracks), and columns `artist`, `track`, `date.entered`, `wk1`, `wk2`, ... `wk76`, for a total of 79 columns. 
The columns `wk1`, `wk2`, ..., `wk76` should indicate that track's rank in that week.

```{r}
billboard_tracks = billboard_weekly %>% 
  pivot_wider(names_from = week, values_from = rank)
```

# Problem 2

All of the following problems purposefully do not specify whether you should use `billboard_weekly` or `billboard_tracks`, or how to get the answer. This is left up to you to figure out. None of the problems rely on each other from here on out

You should answer by starting with one of those dataframes, performing some manipulation, and letting the output print to console, no need to save any output from here on out.

---

**Which track had the best ranking (lowest number) in its very first week in the Billboard Top 100?**

```{r}
# This one can be done quickly with either dataframe
billboard_tracks %>% 
  slice_min(wk1, n = 1)

billboard_weekly %>% 
  filter(week == "wk1") %>% 
  slice_min(rank, n = 1)
```

# Problem 3

Is there any `track` name that got used by two different artists?

```{r}
billboard_tracks %>% 
  count(track) %>% 
  filter(n > 1)

# This one is difficult to do with billboard_weekly unless you know distinct(), which we haven't covered in this course.
billboard_weekly %>% 
  distinct(artist, track) %>% 
  count(track) %>% 
  filter(n > 1)
```

# Problem 4

Three songs that entered the Top 100 in 2000 stayed there all the way to April 7th, 2001.

```{r}
billboard_dates = read_csv("../../data/billboard_2000_dates.csv")

longest_running_songs = slice_max(billboard_dates, date.of.week, n = 3)

longest_running_songs
```

**What were their rankings in that final week?** Your answer should use a `*_join` command to join `longest_running_songs` to EITHER `billboard_weekly` or `billboard_tracks`.

```{r}
# Numerous ways to do this problem using billboard_weekly; here are three possible options

left_join(longest_running_songs, billboard_weekly)

inner_join(longest_running_songs, billboard_weekly)

left_join(billboard_weekly, longest_running_songs) %>% slice_max(date.of.week, n = 3)

left_join(billboard_weekly, longest_running_songs) %>% drop_na(date.of.week)
```

# Problem 5

**Which track had the biggest rise from week 1 to week 5?**

```{r}
# Straightforward with billboard_tracks
billboard_tracks %>%
  mutate(rise = wk1 - wk5) %>% 
  slice_max(rise, n = 1)

# Very difficult to do with billboard_weekly; also relies on the "diff" command, which returns how far apart two numbers are, which we haven't learned in this class
billboard_weekly %>% 
  filter(week %in% c("wk1", "wk5")) %>% 
  group_by(artist, track) %>% 
  summarize(rise = diff(rank)) %>%
  ungroup() %>% 
  slice_max(rise, n = 1)

```


# Problem 6

**Which five tracks spent the most total weeks in the Billboard Top 100?**

```{r}
# This one is very hard to do with billboard_tracks

billboard_weekly %>% 
  drop_na() %>% 
  count(artist, track) %>% 
  slice_max(n, n = 5)
```

# Problem 7

**Which tracks hit rank #1 at ANY point?**

```{r}
# Could do this with billboard_tracks, but would have to do
# filter(wk1 == 1 | wk2 == 1 | wk3 == 1 | ...)

billboard_weekly %>% 
  group_by(artist, track) %>% # If you just group_by(track), "Where I Wanna Be" could be duplicated! Doesn't matter for this question but something to keep in mind
  summarize(bestRank = min(rank, na.rm = T)) %>% 
  filter(bestRank == 1)
```

# Problem 8

**Which track had the most weeks ranked at #1, and how many weeks was it there?**

```{r}
# Also very hard to do with billboard_tracks

billboard_weekly %>% 
  group_by(artist, track) %>% 
  summarize(numWeeksAtOne = sum(rank == 1, na.rm = T)) %>% 
  ungroup() %>% # The grouping of "artist" persists here, have to get rid of it before slice_max
  slice_max(numWeeksAtOne, n = 1)
```

