---
title: "Section 11: Statistical Inference"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
source("../../scripts/ggprob.R")
```

## Confidence intervals

For a 95% CI, let's look at the interval with area 0.95 on the standard normal curve.  The area outside is 0.05, which means each tail has area 0.05/2 = 0.025

```{r}
gnorm(mu = 0, sigma = 1, col = "black") +
  geom_norm_fill(a = -1.96, b = 1.96, fill = "goldenrod") +
      annotate("text", x = 0, y = 0.2, label = "0.95", size = 8) 

gnorm(mu = 0, sigma = 1, col = "black") +
  geom_norm_fill(a = -1.96, b = 1.96, fill = "goldenrod") +
      annotate("text", x = c(0, -2.75, 2.75), y = c(0.2, 0.05, 0.05),
               label = c("0.95", "0.025", "0.025"), size = c(8, 6, 6)) 
```

Find these values with `qnorm`:

```{r}
qnorm(0.025)
qnorm(0.975)
```

What if we wanted to build a 90% CI?

```{r}
# qnorm()
# qnorm()
```


## Hypothesis testing

In the "Lady tasting tea" example, if Bristol is guessing randomly, the number of correct guessses is Binom(8, 0.5).

```{r}
gbinom(8, 0.5, col = "black") +
  geom_binom_density(8, 0.5, a = 8, lwd = 3, col = "dodgerblue")
```

If she gets 6 out of 8 correct, then the p-value is the sum of the binomial probabilities greater than or equal to 6.

```{r}
pbinom(5, 8, 0.5, lower.tail = F)
```

What is the p-value if she gets 8 out of 8 correct?

```{r}
# pbinom()
```








