---
title: "week10-regressionInference"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


```{r}
mendota <- read_csv("~/Desktop/STAT 240/data/lake-mendota.csv")
```

# Sec 006 Monday 11/4


```{r}
model_mendota <- lm(duration~year1, data = mendota)
summary(model_mendota)
```

$95%$ confidence interval for $\hat{\beta}_1$

```{r}
# 95% confidence interval for beta_1_hat

n = length(mendota$year1)
c = 0.95
qcs <- qt(c + (1-c)/2, n-2) # n-2 = length of the data set - 2
qcs

moe <- qcs*0.02735 # moe (margin of error = quantile score*standard error)

c(-0.19445 -  moe, -0.19445 + moe)
```



```{r}
c = 0.99
qcs <- qt(c + (1-c)/2, n-2) # n-2 = length of the data set - 2
qcs

moe <- qcs*0.02735 # moe (margin of error = quantile score*standard error)

c(-0.19445 -  moe, -0.19445 + moe)
```


# Sec 003 Monday 11/4

```{r}
mendota <- read_csv("~/Desktop/STAT 240/data/lake-mendota.csv")
```

```{r}
model_mendota <- lm(duration~year1, data = mendota)
summary(model_mendota)
```

Build a 95% confince interval for $\hat{\beta}_1$

```{r}
n = length(mendota$year1)
c = 0.95
qcs = qt(c + (1-c)/2, n-2) # critical values, and n is the length of the sample 
qcs

moe = qcs*0.02735 # moe is the margin of error (= critical value * standard error)

c(-0.19445 - moe, -0.19445 + moe)
```


```{r}
c = 0.99
qcs = qt(c + (1-c)/2, n-2) # critical values, and n is the length of the sample 
qcs

moe = qcs*0.02735 # moe is the margin of error (= critical value * standard error)

c(-0.19445 - moe, -0.19445 + moe)
```

# Sec 005 Monday Nov 4

```{r}
mendota <- read_csv("~/Desktop/STAT 240/data/lake-mendota.csv")
```
```{r}
model_mendota <- lm(duration~year1, mendota)
summary(model_mendota)
```
A 95% confidence interval for $\hat{\beta}_1$

```{r}
n <- length(mendota$year1)
c <- 0.95

cv <- qt(c + (1-c)/2, n-2) # quantile confidence score
cv

moe <- cv*0.02735 # moe is the margin of error (= critical value * Standard error) 
moe

c(-0.19445 - moe, -0.19445 + moe)
```

99% confidence interval

```{r}
n <- length(mendota$year1)
c <- 0.99

cv <- qt(c + (1-c)/2, n-2)
cv

moe <- cv*0.02735
moe

c(-0.19445 - moe, -0.19445 + moe)
```

# Sec 006 Wednesday Nov 6

```{r}
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm")
summary(model_mendota)
```

- Linearity of residuals ($\epsilon_i$) - satisfied
- Normality $\epsilon_i$ around 0 - satisfied
- Constant variance of $\epsilon_i$ - satisfied

```{r}
residual <- resid(model_mendota)
ggplot(mapping = aes(x = mendota$year1, y = residual)) +
  geom_point() +
  geom_hline(yintercept = 0) 
```

$$H_0 : \beta_1 = 0, H_{\alpha} < 0$$

- Third/Fourth step (calculating the test statistic)

$$t = \frac{\hat{\beta}_1 - \beta_{1,H_0}}{s.e(\beta_1)}$$

```{r}
t =  -0.19445 / 0.02735
t
```

Step 5: Calculating the p-value

```{r}
p_value = pt(q = -7.109689, df = n-2)
p_value 
```

Step 6: $\alpha = 0.05, 0.01$

Since p_value < $\alpha$, we reject $H_0$, or we can say we have enough evidence to say that $\beta_1<0$.



```{r}
y_pred <- 478.86548 - 0.19445*2024
y_pred
```


# Sec 003 Wednesday 11/06

```{r}
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm")
summary(model_mendota)
```

Step 1: Checking the model

```{r}
residual <- resid(model_mendota)
ggplot(mapping = aes(x = mendota$year1, y = residual)) +
  geom_point() +
  geom_hline(yintercept = 0)
```

- Linearity of residual ($\epsilon_i$) -> satisfied
- Normality around 0 of $\epsilon_i$ -> satisfied
- Constant deviation or variance of $\epsilon_i$ -> satisfied

Step 2:

$$H_0: \beta_1 = 0, H_{\alpha}: \beta_1 < 0$$
Step 3: test statistic 

$$t = \frac{\hat{\beta}_1}{s.e(\hat{\beta_1})}$$

Step 4: Calculate the test statistics

```{r}
summary(model_mendota)
t = -0.19445/0.02735
t
```
Step 5: Calculate the p-value

```{r}
pt(t, n-2)
```

Step 6: $\alpha=0.05$

p_value < $\alpha$

We reject $H_0$. We have enough evidence to support that $\beta_1 < 0$.


95% prediction interval values

```{r}
prediction_interval <- predict(model_mendota, newdata = tibble(year1 = 1850:2023), interval = "predict")
prediction_interval

confidence_interval <- predict(model_mendota, newdata = tibble(year1 = 1850:2023), interval = "confidence")
confidence_interval

```


```{r}
ggplot(mendota, aes(year1, duration)) +
  geom_point() +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = upr))

ggplot(mendota, aes(year1, duration)) +
  geom_point() +
  geom_line(data = as_tibble(confidence_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(confidence_interval), mapping = aes(x = 1850:2023, y = upr)) +
  geom_smooth(method = "lm") +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = upr))
```
# Sec 005 Wednesday 11/06

```{r}
ggplot(mendota, aes(x = year1, y = duration)) +
  geom_point() +
  geom_smooth(method = "lm")
```

Step 1: Model




```{r}
residual <- resid(model_mendota)
ggplot(mapping = aes(x = mendota$year1, y = residual)) +
  geom_point() +
  geom_hline(yintercept = 0)
```
- Linearity of the residuals ($\epsilon_i$) <- satisfied
- Normality of $\epsilon_i$ around 0 <- satisfied
- Constant variance/deviation of $\epsilon_i$ <- satisfied

Step 2: Construct youy hypothesis

$$H_0: \beta_1 = 0, H_{\alpha}: \beta_1 <0$$


Step 3: identify the test statistic and null distribution:

$$t = \frac{\hat{\beta}_1 - 0}{s.e(\hat{\beta}_1)} \sim t(n-2)$$

Step 4: Calculating the test statistic


```{r}
summary(model_mendota)
t <- -0.19445/0.02735
t
```
Step 5: Calculate the p_value

```{r}
p_value <- pt(t, n-2)
p_value
```

Step 6: Comparing with $\alpha = 0.05$

p_value < 0.05, so we have enough evidence to reject our null hypothesis or support the alternative hypothesis

> Prediction interval

```{r}
prediction_interval <- predict(model_mendota, newdata = tibble(year1 = 1850:2023), interval = "prediction")

ggplot(mendota, aes(year1, duration)) +
  geom_point() +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = upr))
```
Confidence interval for E(Y|X)

```{r}
confidence_interval <- predict(model_mendota, newdata = tibble(year1 = 1850:2023), interval = "confidence")

ggplot(mendota, aes(year1, duration)) +
  geom_point() +
  geom_line(data = as_tibble(confidence_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(confidence_interval), mapping = aes(x = 1850:2023, y = upr)) +
  geom_smooth(method = "lm") +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = lwr)) +
  geom_line(data = as_tibble(prediction_interval), mapping = aes(x = 1850:2023, y = upr))
```





















