---
title: "week12-inferenceMeans"
output: html_document
date: "2024-11-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
bm_orig = read_table("~/Desktop/STAT 240/data/TIM.txt",
                col_types = cols(.default = col_double()))

bm = bm_orig %>% 
  filter(!is.na(`K40-Fin`)) %>% 
  filter(Year < 2013) %>% 
  arrange(Year, BibNum) %>% 
  select(-starts_with("Start"), -HalfMar, -Age2014) %>% 
  mutate(Sex = case_when(
    Gender1F2M == 1 ~ "female",
    Gender1F2M == 2 ~ "male")) %>% 
  mutate(Time = pmap_dbl(select(., starts_with("K")), sum)) %>% 
  mutate(age1 = case_when(
    Age < 35 ~ 18,
    Age >= 80 ~ 80,
    TRUE ~ floor(Age/5)*5),
    age2 = case_when(
      Age < 35 ~ 34,
      TRUE ~ age1 + 4),
    Age_Range = case_when(
      age1 == 80 ~ "80 and older",
      TRUE ~ str_c(age1,"-",age2))) %>% 
  select(-Gender1F2M, -age1, -age2) %>% 
  relocate(BibNum, Year, Sex, Age, Age_Range, Time)

bm
```


# Sec 006 Monday 11/18

```{r}
bm_summary = bm %>% 
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  summarize(averageTime = mean(Time), sdTime = sd(Time), n = n())

bm_summary
```

```{r}
x_bar <- 236
s <- 36.2  
n <- 3557
```

- 95% confidence interval for $\bar{x}$,

$$\bar{x} \pm qt(C+(1-C)/2, n-1)*\frac{s}{\sqrt{n}}$$

```{r}
c = 0.95
cv <- qt(c + (1-c)/2, n-1)
cv
s.e <- s/sqrt(n)

x_bar + c(-1,1)*cv*s.e
```

# Sec 003 Monday 11/18

```{r}
bm_summary = bm %>% 
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  summarize(averageTime = mean(Time), sdTime = sd(Time), n = n())

bm_summary
```

- 95% confidence interval for $\bar{x}$,

$$\bar{x} \pm qt(C+(1-C)/2, n-1)*\frac{s}{\sqrt{n}}$$

```{r}
x_bar <- 235.5062
s <- 36.2
n <- 3557

s.e <- s/sqrt(n)
s.e

cv <- qt(0.975, n-1)
cv

#qnorm(0.975)

x_bar + c(-1,1)*cv*s.e
```


# Sec005 Monday 11/18

95% Confidence Interval: 

$$(\bar{x} - qt(c+(1-c)/2, n-1)*\frac{s}{\sqrt{n}}, \bar{x} - qt(c+(1-c)/2, n-1)*\frac{s}{\sqrt{n}})$$


```{r}
bm_summary = bm %>% 
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  summarize(averageTime = mean(Time), sdTime = sd(Time), n = n())

bm_summary
```

```{r}
n <- 3557
x_bar <- 235.5062
s <- 36.2

s.e <- s/sqrt(n)
s.e
cv <- qt(0.975, n-1)
cv

# qnorm(0.975)

x_bar + c(-1,1)*cv*s.e
```


# Sec 006 Wednesday 11/20 

```{r}
times = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% pull(Time)

#times = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% select(Time)

times

t.test(times, conf.level = 0.99) 
```
## Hypothesis Testing for single mean

- Model Statement:

$$X \sim D(\mu, \sigma)$$
- State the hypothesis

$$H_0: \mu = 240, H_{\alpha} \neq 240$$
- Test statistic and the null distribution

$$t = \frac{\bar{x}-\mu_{Null}}{s/\sqrt{n}} \sim t(n-1) $$
- Calculate the values from the dataset


```{r}
x_bar <- 235.5062	
s <- 36.1675
n <- 3557
t <- (x_bar- 240)/(s/sqrt(n))
t
```

- Calculate the p-value

```{r}
pt(t,n-1)
p_val <- 2*pt(t,n-1)
p_val
```

- Comparing the p-val with $\alpha = 0.05$

We reject the null hypothesis, that is the true average of female runners in the age group 18-34 is different from 240 

```{r}
t.test(times, mu = 240, alternative = "two.sided")
```

```{r}
t.test(times, mu = 240, alternative = "less")
```

```{r}
bm_twomeans_summary = bm %>% 
  filter(Sex == "female", Age_Range == "18-34") %>% 
  group_by(Year) %>% 
  summarize(avgTime = mean(Time), sdTime = sd(Time), n = n())

bm_twomeans_summary
```

```{r}
x_bar_1 <- 235.5062	
x_bar_2 <- 235.2337
s1 <- 36.16750	
s2 <- 36.59875	
n1 <- 3557
n2 <- 3580

w <- (s1^2/n1 + s2^2/n2)^2 / (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1))) # degrees of freedom
w

cv <- qt(0.975, w)
cv

(x_bar_1 - x_bar_2) + c(-1,1)*cv*sqrt((s1)^2/n1 + (s2)^2/n2)
```


```{r}
times_2010 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% pull(Time)

times_2011 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2011) %>% pull(Time)

t.test(times_2010, times_2011)
```

$$H_0: \mu_1 - \mu_2 =0, H_{\alpha}: \mu_1-\mu_2 > 0 $$

# Sec 003 Wednesday 11/20

```{r}
times = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% pull(Time)

#times = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% select(Time)

times

t.test(times)

t.test(times, conf.level = 0.80)

?t.test
```
## Hypothesis Testing

- Step 1: Model Statement

$$X \sim D(\mu, \sigma)$$
$\mu$ is the true average time taken to complete the boston marathon by a female participant in the age group of 18-34.

Assumption: The data points are independent of each other

- Step 2: State the hypothesis:


$$H_0: \mu = 240, H_{\alpha}: \mu \neq 240 $$
- Step 3: Test statistic and the null distribution

$$t = \frac{\bar{x}- \mu_{null}}{s.e(\bar{x})} = \frac{\bar{x}- \mu_{null}}{s/\sqrt{n}} \sim t(n-1) $$

- Step 4: Evidence from the data set for the alternative hypothesis

```{r}
x_bar <- 235.5062	
s <- 36.1675	
n <- 3557

t = (x_bar - 240)/(s/sqrt(n))
t
```


- Step 5: Calculate the p-value

```{r}
pt(t, n-1)

p_val <- 2*pt(t, n-1)
p_val
```

- Step 6: Compare p-value with $\alpha$

$$\alpha = 0.05$$

Since p-value is less than $\alpha$ we reject $H_0$.


```{r}
t.test(times, mu = 240)

t.test(times, mu = 240, alternative = "less")
```

## Confidence interval for difference of means

$$(\bar{x}_1-\bar{x}_2) \pm qt(C+(1-c)/2, W)*\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}$$

```{r}
times_2010 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% pull(Time)

times_2011 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2011) %>% pull(Time)

t.test(times_2010, times_2011)
```
```{r}
x_bar_1 <- 235.5062	
x_bar_2 <- 235.2337
s1 <- 36.16750	
s2 <- 36.59875	
n1 <- 3557
n2 <- 3580

w <- (s1^2/n1 + s2^2/n2)^2 / (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))
cv <- qt(0.975, w)
cv

(x_bar_1 - x_bar_2) + c(-1,1)*cv*sqrt((s1)^2/n1 + (s2)^2/n2)
```


# Sec 005 Wednesday 11/20

```{r}
times = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% select(Time)

times

t.test(times)

?t.test

t.test(times, conf.level = 0.80)
```
## Hypothesis testing for the single mean

- Step 1: Model Statement:

$$X \sim D(\mu, \sigma)$$
Assumption: Assume that time taken by individual participants is independent of each other.

- Step 2: State hypothesis:

$$H_0: \mu = 240, H_{\alpha}: \mu \neq 240 $$
- Step 3: Identify test statistic and null distribution

$$t = \frac{\bar{x}-\mu_{null}}{s.e(\bar{x})} = \frac{\bar{x}-\mu_{null}}{s/\sqrt{n}} \sim t(n-1) $$

- Step 4: Evidence for the alternative hypothesis from the dataset

```{r}
x_bar <- 235.5062	
s <- 36.1675	
n <- 3557

t = (x_bar - 240)/(s/sqrt(n))
t

```

- Step 5: Calculate p-value

```{r}
p_val <- 2*pt(t, n-1) # we multiply by two because this is a two sided test
p_val
```

- Step 6: Conclusion depending on $\alpha$

$$\alpha = 0.05$$

Since p-value is les than alpha, we reject $H_0$. We have enough evidence that $\mu$ is not equal to 240, in fact it is less than 240.

```{r}
t.test(times, mu = 240, alternative = "less")
t.test(times, mu = 240)
```

## Confidence interval for $\mu_1 - \mu_2$


```{r}
x_bar_1 <- 235.5062	
x_bar_2 <- 235.2337
s1 <- 36.16750	
s2 <- 36.59875	
n1 <- 3557
n2 <- 3580

w = (s1^2/n1 + s2^2/n2)^2 / (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))
w

cv <- qt(0.975, w)
cv

(x_bar_1 - x_bar_2 ) + c(-1, 1)*cv*sqrt((s1)^2/n1 + (s2)^2/n2)
```

```{r}
times_2010 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2010) %>% pull(Time)

times_2011 = bm %>% filter(Sex == "female" & Age_Range == "18-34" & Year == 2011) %>% pull(Time)

t.test(times_2010, times_2011)
```
## Hypothesis testing for difference of means

- Step 1: $$X_1 \sim D_1(\mu_1, \sigma_1), X_2 \sim D_2(\mu_2, \sigma_2)$$

Assumption: All the observations of $X_1$ and $X_2$ are independent.


- Step 2: State the hypothesis

$$H_0: \mu_1=\mu_2, H_{\alpha}: \mu_1 \neq \mu_2  $$

- Step 3: Identify test statistic and null distribution



# Sec 006 Friday 11/22

- Step 1: Model Statement:

$$X \sim D_x (\mu_x, \sigma_x)$$
where, $\mu_x$ is the true average time taken by the female participants in the age group 18-34 in 2010 Boston marathon 

$$Y \sim D_y(\mu_y, \sigma_y)$$
where, $\mu_y$ is the true average time taken by the female participants in the age group 18-34 in 2011 Boston marathon 

Assumption: The variable $X$ and $Y$ are independent 

- Step 2: State the hypothesis

$$H_0: \mu_1 - \mu_2 =0, H_{\alpha}: \mu_1-\mu_2 \neq 0 $$

- Step 3: Identify test statistic and the null distribution

$$t = \frac{(\bar{x}-\bar{y})- (\mu_x - \mu_y)}{\sqrt{\frac{s_x^2}{n_x}+\frac{s_y^2}{n_y}}} \sim t(n-1) $$

- Step 4: Calculate the test statistic

```{r}
bm_twomeans_summary = bm %>% 
  filter(Sex == "female", Age_Range == "18-34") %>% 
  group_by(Year) %>% 
  summarize(avgTime = mean(Time), sdTime = sd(Time), n = n())

bm_twomeans_summary


x_bar_1 <- 235.5062	
x_bar_2 <- 235.2337
s1 <- 36.16750	
s2 <- 36.59875	
n1 <- 3557
n2 <- 3580

t = (x_bar_1 - x_bar_2)/(sqrt(s1^2/n1 + s2^2/n2))
t
```

- Step 5: Calculate p- value

```{r}
w = (s1^2/n1 + s2^2/n2)^2 / (s1^4/(n1^2*(n1-1)) + s2^4/(n2^2*(n2-1)))
w

p_val <- 2*(1 - pt(t,w))
p_val
```


- Step 6: Compare with $\alpha$ and make a conclusion

$$\alpha = 0.05$$

We have enough evidence that the true average time taken by the female (18-34) participants of 2010 and 2011 Boston marathon is the same.

```{r}
times_2010 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  pull(Time)

times_2011 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2011) %>% 
  pull(Time)

t.test(times_2010, times_2011)
```

## Paired t.test()


```{r}
thirteen = c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen = c(46.3, 60.5, 68.2, 59.4, 50.6)
```


We want to compare the mean of the heights at 13 and 14. Since the variables are dependent, we consider doing the paired t-test

```{r}
growth <- fourteen - thirteen
growth

t.test(growth)
t.test(fourteen-thirteen, alternative = "greater")
```

```{r}
t.test(fourteen, thirteen) # for difference of means which is an unpaired t.test.

t.test(fourteen, thirteen, paired = TRUE)
```


# Sec 003 Friday 11/22

- Step 1: Model statement

$$X \sim D_x(\mu_x, \sigma_x)$$
$$Y \sim D_y(\mu_y, \sigma_y)$$

where, $\mu_x$ and $\mu_y$ is the true average time taken by the female (18-34) participants in 2010 and 2011, resp. 


Assumption: $X$ and $Y$ are independent random variables.

- Step 2: State the hypothesis

$$H_0: \mu_x = \mu_y \implies \mu_x - \mu_y = 0 \\
H_{\alpha}: \mu_x \neq \mu_y \implies \mu_x-\mu_y \neq 0$$

- Step 3: Identify test statistic and null distribution

- Step 4: Calculate the test statistic for the data

```{r}
x_bar <- 235.5062	
y_bar <- 235.2337
sx <- 36.16750	
sy <- 36.59875	
nx <- 3557
ny <- 3580

t = (x_bar - y_bar)/(sqrt(sx^2/nx + sy^2/ny))
t 
```


- Step 5: Calculate the p-value

```{r}
w = (sx^2/nx + sy^2/ny)^2 / (sx^4/(nx^2*(nx-1)) + sy^4/(ny^2*(ny-1)))
w

p_val <- 2*pt(- abs(t), w)
2*(1- pt(t, w))

p_val
```

- Step 6: Compare $\alpha = 0.05$

```{r}
times_2010 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  pull(Time)

times_2011 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2011) %>% 
  pull(Time)
t.test(times_2010, times_2011)
```



```{r}
thirteen = c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen = c(46.3, 60.5, 68.2, 59.4, 50.6)
t.test(fourteen, thirteen, alternative = "less")
```
```{r}
growth <- thirteen - fourteen
t.test(growth, alternative = "less")
t.test(thirteen-fourteen, alternative = "less")
t.test(thirteen, fourteen, paired = T, alternative = "less")
```

# Sec 005 Friday 11/22

- Step 1: Model Statement:

$$X = \text{time taken by a female participant in 2010 Bm} \sim D_x(\mu_x, \sigma_x)$$
$$Y = \text{time taken by a female participant in 2011 Bm} \sim D_y(\mu_y, \sigma_y)$$
where, $\mu_x$ and $\mu_y$ are the true average time taken by female participants in 2010 and 2011 Boston marathon, resp.

Assumption: $X$ and $Y$ are independent rv

- Step 2: State the hypothesis

$$H_0: \mu_x = \mu_y \implies \mu_x-\mu_y = 0 \\
H_{\alpha}:  \mu_x \neq \mu_y \implies \mu_x-\mu_y \neq 0$$


- Step 3: Identify the test statistic and the null distribution  

- Step 4: Calculate the test statitic from the data set

```{r}
x_bar <- 235.5062	
y_bar <- 235.2337
sx <- 36.16750	
sy <- 36.59875	
nx <- 3557
ny <- 3580

t = (x_bar - y_bar)/(sqrt(sx^2/nx + sy^2/ny))
t
```


- Step 5: Calculate the p-value

```{r}
w = (sx^2/nx + sy^2/ny)^2 / (sx^4/(nx^2*(nx-1)) + sy^4/(ny^2*(ny-1)))

p_value <- 2*pt(-abs(t), w) 
p_value

2*(1 - pt(t, w))
```


- Step 6: Compare p-value with $\alpha = 0.05$

We fail to reject that there is no difference in the true average time taken by female participants in the age group of 18-34 from the 2010 and 2011 Boston marathon  

```{r}
times_2010 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  pull(Time)

times_2011 = bm %>%
  filter(Sex == "female", Age_Range == "18-34", Year == 2011) %>% 
  pull(Time)

t.test(times_2010, times_2011)
```
```{r}
thirteen = c(44.1, 59.0, 65.9, 58.7, 49.3)
fourteen = c(46.3, 60.5, 68.2, 59.4, 50.6)
t.test(thirteen, fourteen, alternative = "greater")
```


```{r}
z <- fourteen - thirteen
t.test(z, alternative = "greater")
t.test(fourteen-thirteen, alternative = "greater")
t.test(fourteen, thirteen, paired = T, alternative = "greater")
```












