---
title: "Discussion 1"
output: html_document
---

```{r setup, include=FALSE}
# We will usually provide a chunk like this for you.
# This is setting "global chunk options"; "global" meaning applying to every chunk in the rest of the document.
#   echo = TRUE causes every chunk to appear in the knitted file.
#   warning = FALSE causes warning messages (different than errors) to NOT appear in the knitted file.
#   message = FALSE causes other messages (like those from loading packages) to NOT appear in the knitted file.

# Notice include = FALSE as a chunk option at the top of this chunk in line 6. This means that neither this chunk nor its output will appear in the knitted file.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Preliminaries

- This file should be in `STAT240/discussion/ds01`.

# Problem 1

**What is the date for Exam 1 of this course? Is it take-home or in-person?**

> Replace this text with your response to problem 1.

# Problem 2

**What is the date for Exam 2 of this course? Is it take-home or in-person?**

> Replace this text with your response to problem 2.

# Problem 3

**Use the following R chunk to compute three squared plus five.**

You may find [Section 2.3.2 of the textbook helpful.](https://bwu62.github.io/stat240-revamp/rstudio-intro.html#r-math)

```{r}
# Write your code in here!

```

Please ensure that the answer is appearing in the knitted file correctly. Do not just write 14.

# Submission

When you have answered the above questions, please knit this file and submit **BOTH** the R Markdown file (.Rmd) and the knitted html file (.html).

We will ask for both files for every assignment in this course.

We will grade directly from the .html file, so make sure you are satisfied with how it looks before you submit. We ask for your .Rmd as a backup in case your .html file has errors in it. 