---
title: "week07-binomial and normal"
output: html_document
date: "2024-10-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../../scripts/ggprob.R")
```

# Sec 006 10/16

```{r}
# For 10 coin flips. You want to find the probability of getting 4 heads

dbinom(4, 10, 0.5)

```

```{r}
# For 1 coin flips. You want to find the probability of getting 1 heads

dbinom(1, 1, 0.5)
```

# Sec 003 10/16

```{r}
dbinom(0:10, 10, 0.5)

dbinom(1, 1, 0.5)
```

```{r}
set.seed(29900)
rbinom(100, 10, 0.5)

rbinom(100, 1, 0.5)

rbinom(100, 2, 0.5)
```

# Sec 005 10/16

```{r}
# probability of getting 2 head when a coin is flipped 10 times

dbinom(2, 10, 0.5) # using the dbinom function

choose(10, 2) * (0.5)^2 * (0.5)^8 # using the formula

dbinom(1, 1, 0.5)
```

# Sec 006 10/18

```{r}
# Create a data set with 100 observations for 10 coin flips (X will be the number of heads)

data = rbinom(1000, 10, 0.5)
mean_s = mean(data)
variance_s = var(data)
mean_s
variance_s 


# Find the mean of Binom(10, 0.5)

mean_d = 10*0.5
variance_d = 10*0.5*0.5
mean_d 
variance_d

dbinom(8, 10, 0.5)
mean(data==8)
```

Graph for binomial distribution

```{r}
gbinom(10, 0.5)
```

```{r}
# Probability of getting atmost 3 heads

pbinom(3, 10, 0.5) # gives you the added up prob to the left of x

# Probability of getting atleast 3 heads

pbinom(3, 10, 0.5, lower.tail = FALSE) # gives you the added up prob to the right of x

1 - pbinom(3, 10, 0.5)
```

```{r}
# The number of heads such that it is above 70% of the usual number of heads

qbinom(0.7, 10, 0.5) # an approximation of the smallest number such that that prob is reached

pbinom(6, 10, 0.5)
pbinom(5, 10, 0.5)
```

# Sec 003 10/18

```{r}
# Data for flipping a rock which has a prob of 20% of landing face up. repeat the experiment 100 times where you flip the rock 10 times to see how many times it lands face up.

set.seed(2400)
data <- rbinom(100, 10, 0.2) 

data

mean(data) 
var(data)
```

```{r}
# You are flipping a coin 10 times. Find the probability of getting atmost 3 heads

pbinom(3, 10, 0.5) # P(X <= 3)

dbinom(0, 10, 0.5) + dbinom(1, 10, 0.5) + dbinom(2, 10, 0.5) + dbinom(3, 10, 0.5) 

sum(dbinom(0:3, 10, 0.5))
```

```{r}
# You are flipping a coin 10 times. Find the probability of getting atleast 3 heads

pbinom(3, 10, 0.5, lower.tail = FALSE)

1 - pbinom(3, 10, 0.5)

sum(dbinom(4:10, 10, 0.5))
```

```{r}
# Graph the distribution of Binom(10, 0.2)

gbinom(10, 0.2) # this is not an inbuilt r function. So you have to use the source ar ggprob.R 

ggplot(tibble(x=data), aes(x = x)) + geom_histogram()
```

```{r}
# You are flipping a coin 10 times. Find the number of heads such that it is above 70% of the usual number of heads



qbinom(0.7, 10, 0.5) # this gives an approx or least possible value such that P(X<= x) = 0.7

pbinom(6, 10, 0.5)
pbinom(5, 10, 0.5)
```

# Sec 005 10/18

mean: $E(X)=\mu = n\cdot p$

variance: $\sigma^2 = n \cdot p \cdot (1-p)$

```{r}
data <- rbinom(100, 10, 0.5)
data

mean(data)
var(data)

n = 10
p = 0.5

mean_dist = n*p
mean_dist

variance_dist = n*p*(1-p)
variance_dist
```

-   `dbinom(x, size, prob)` where d = density, gives $P(X=x)$
-   `pbinom(x, size, prob)` where p = probability, gives p such that $P(X\leq x)= p$
-   `rbinom(n, size, prob)` where r = random, generates random points from the binom(size, prob)
-   `qbinom(p, size, prob)` where q = quantile, gives approx or the smallest value of `x` such that $P(X\leq x) = p$

```{r}
# You are flipping a coin 10 times. Find the probability of getting atmost 3 heads

pbinom(3, 10, 0.5) # P(X <= 3)
dbinom(0, 10, 0.5) + dbinom(1, 10, 0.5) + dbinom(2, 10, 0.5) +dbinom(3, 10, 0.5)
sum(dbinom(0:3, 10, 0.5))
```

```{r}
# You are flipping a coin 10 times. Find the probability of getting atleast 3 heads

pbinom(3, 10, 0.5, lower.tail = FALSE)
1 - pbinom(3, 10, 0.5)
```

```{r}
gbinom(10, 0.5)

ggplot(data = tibble(x = rbinom(1000, 10, 0.5)), aes(x = x)) + geom_histogram()
```

```{r}
# You are flipping a coin 10 times. Find the number of heads such that it is above 70% of the usual number of heads

qbinom(0.7, 10, 0.5) # approx or it gives the smallest value of x such that P(X<= x) = 0.7 

pbinom(6, 10, 0.5)
pbinom(5, 10, 0.5)
```



## Sec 006 Monday 10/21

Find the density at the point $x=16$ when $X \sim N(10, 6)$.

```{r}
dnorm(16, 10, 6)
```

Find the area to the right and left of $x=16$ when $X \sim N(10, 6)$.

```{r}
pnorm(16, 10, 6) # prob to the right of 16
1 - pnorm(16, 10, 6) 
pnorm(16, 10, 6, lower.tail = F)
```

```{r}
gnorm(10, 6)+
  geom_norm_density(10, 6, a = NULL, b = 16, color = "red") +
  geom_norm_fill(10, 6, a = NULL, b = 16, color = "red") +
  geom_vline(xintercept = 10+c(-2, 2)*6)
```

```{r}
pbinom(4, 10, 0.5)
pbinom(4.1, 10, 0.5)

pbinom(5, 10, 0.5)
pbinom(6, 10, 0.5)

pnorm(4, 10, 5)
pnorm(4.1, 10, 5)

pnorm(6, 10, 5)
```

```{r}
ggplot() +
  geom_norm_density(10, 5, color = "red") +
  geom_norm_density(10, 3, color = "blue") +
  geom_norm_density(10, 10, color = "green")
```

# Sec 003 Monday 10/21


```{r}
gbinom(10, 0.5) +
  geom_binom_density(10, 0.5, a = NULL, b = 4, color = "red") +
  geom_binom_density(10, 0.5, a = 8, b = NULL, color = "green")
```

Normal Distribution:

- density at a point (P(X=x)): dnorm() 

```{r}
dnorm(16, 10, 6)
```

- probability to the left of a point x $(P(X\leq x))$: pnorm()

```{r}
pnorm(16, 10, 6) # P(X <= 16)
```

- probability to the right of a point x $(P(X\geq x))$: 1 - pnorm(), pnorm(..., lower.tail = F)

```{r}
1 - pnorm(16, 10, 6)
pnorm(16, 10, 6, lower.tail = F)
```

- given q, the smallest x value such that P(X <= x) = q

```{r}
qnorm(0.5, 10, 6) # gives the mean since in normal distribution the mean lies at the center of the distribution
```

```{r}
ggplot() +
  geom_norm_density(10, 3, color = "red") +
  geom_norm_density(10, 5, color = "blue") +
  geom_norm_density(10, 7, color = "green")
```

# Sec 005 Monday 10/21

```{r}
gbinom(10, 0.5) +
  geom_binom_density(10, 0.5, a = NULL, b = 4, color = "red") +
  geom_binom_density(10, 0.5, a = 8, b = NULL, color = "green")
```


```{r}
gnorm(10, 6) # graph of the N(10, 6)
```


```{r}
ggplot() +
  geom_norm_density(10, 3, color = "red") +
  geom_norm_density(10, 5, color = "blue") +
  geom_norm_density(10, 7, color = "green")
```

- density at a point x, P(X = x): dnorm()

```{r}
dnorm(16, 10, 6) # this f(16) or an approximation of P(X=16)
```

- probability to the left of a point x, $P(X \leq x)$: pnorm()

```{r}
pnorm(16, 10, 6) # area to the left of 16

gnorm(10, 6) +
  geom_norm_fill(10, 6, b = 16, fill = "blue")
```
- probability to the right of a point x, $P(X \geq x)$, 1 - pnorm(), pnorm(..., lower.tail = FALSE)

```{r}
1 - pnorm(16, 10, 6) # area to the right of 16
pnorm(16, 10, 6, lower.tail = F)
```


```{r}
pnorm(16, 10, 6)
pnorm(16.0001, 10, 6)
```


- given a prob p, gives an approx value of x, such that $P(X \leq x) = p$: qnorm()

```{r}
qnorm(0.5, 10, 6)
qnorm(0.45, 10, 6)
pnorm(9.24, 10, 6)
```

- generates random values from the distribution $N(\mu, \sigma)$: rnorm()

```{r}
rnorm(10, 10, 6)
```


```{r}
x = 16
mu = 10
sigma = 6

z = (x-mu)/sigma
z

pnorm(16, 10, 6)
pnorm(z, 0, 1)
```


# Sec 006 Wednesday 10/23


```{r}
x = 85.1
mu = 80
sigma = 5.1
z = (x - mu)/sigma
z

pnorm(85.1, 80, 5.1)

gnorm(80, 5.1) +
  geom_norm_fill(80, 5.1, b = 85.1, color = "red")


```

```{r}
gnorm(0, 1) +
  geom_norm_fill(0, 1, b = 1, color = "red")
pnorm(1, 0, 1)
```

```{r}
random_binom = tibble(x =  replicate(n = 10000, mean(rbinom(10, 10, 0.5))))

head(random_binom)

gbinom(10, 0.5) +
  geom_density(data = random_binom, aes(x))

random_binom_summarized = random_binom %>% summarise(mean = mean(x), sd = sd(x))
random_binom_summarized
```
```{r}
random_exp = tibble(x =  replicate(n = 100000, mean(rexp(1000, 0.7))))

ggplot() +
  geom_density(data = random_exp, aes(x))
```

# Sec 003 - Wednesday 10/23


```{r}
random_binom = tibble(x = replicate(10000, mean(rbinom(10, 10, 0.5))))

head(random_binom, 20)

gbinom(10, 0.5) +
  geom_density(data = random_binom, aes(x))

random_binom %>% summarise(mean = mean(x), sd = sd(x))
```

```{r}
random_unif = tibble(x = replicate(10000, mean(runif(100, 2, 5))))

head(random_unif)

ggplot() +
  geom_density(data = random_unif, aes(x))
```


# Sec 005 Wednesday 10/23

```{r}
random_binom = tibble(x = replicate(10000, mean(rbinom(10, 10, 0.5))))

head(random_binom, 20)

gbinom(10, 0.5) +
  geom_density(data = random_binom, aes(x), color = "red")
```

```{r}
random_unif = tibble(x = replicate(100000, mean(runif(100, 2, 5))))

head(random_unif, 20)

ggplot() +
  geom_density(data = random_unif, aes(x), color = "red")
```

```{r}
n = 100
p = 0.98
mu = n*p
sigma = sqrt(n*p*(1-p))

gbinom(n, p, scale=TRUE) +
  geom_norm_density(mu, sigma, color = "red") +
  theme_minimal()
```






















