---
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE,
                      warning = FALSE,
                      error = TRUE, fig.height = 3,
                      fig.width = 4)

library(tidyverse)
source("../../scripts/ggprob.R")
```

### dbinom()

- Probability of 2 heads in 5 coins tosses of a fair coin

```{r}
dbinom(2, 5, 0.5)

choose(5, 2) * (0.5)^2 * (0.5)^3
```

- all the probabilities

```{r}
dbinom(0:5, 5, 0.5)

prob1 = tibble(
  x = 0:5,
  p = dbinom(x, 5, 0.5)
)

prob1
```

- sums of probabilities

> If I toss a fair coin 100 times, what is the probability of getting 60 or more heads?

```{r}
sum(dbinom(60:100, 100, 0.5))
```

### Mean and Variance

- $n = 100$, $p = 0.5$

```{r}
n = 100
p = 0.5

mu = n*p
mu

sigma = sqrt( n * p * (1-p) )
sigma
```

### Simulation

```{r}
B = 10^6
n = 100
p = 0.2

prob2 = tibble(
  x = rbinom(B, n, p)
)

ggplot(prob2, aes(x = x)) +
  geom_bar()

prob2 %>% 
  summarize(mean = mean(x),
            sd = sd(x))

n*p
sqrt(n*p*(1-p))
```

### gprob examples

```{r}
ggplot() +
  geom_binom_density(n = 100, p = 0.5, a = 40, b = 55) +
  geom_binom_density(n = 100, p = 0.6, a = 50, b = 70, color = "red", alpha = 0.5)
```

### pbinom()

- What is the probability that there are 60 or more heads in 100 tosses of a fair coin?

- $P(X \ge 60) = 1 - P(X < 60) = 1 - P(X \le 59)$

```{r}
sum( dbinom(60:100, 100, 0.5) )

1 - pbinom(59, 100, 0.5)

gbinom(100, 0.5, scale = T) +
  geom_binom_density(100, 0.5,
                     a = 60,
                     b = 75, color = "red")
```

- $P(35 < X < 55)$

```{r}
sum( dbinom(36:54, 100, 0.5) )

pbinom(54, 100, 0.5) - pbinom(35, 100, 0.5)
```

- $P(35 \le X \le 55)$

```{r}
sum( dbinom(35:55, 100, 0.5) )

pbinom(55, 100, 0.5) - pbinom(34, 100, 0.5)
```

### qbinom()

$X \sim \text{Binomial}(100, 0.5)$

What is the 0.4 quantile of this distribution?

(What is the 40th percentile?)

Seeking a value $x$ such that $P(X \le x) \ge 0.40$ and $P(X \ge x) \ge 0.6$.

```{r}
qbinom(0.4, 100, 0.5)

pbinom(49, 100, 0.5)

pbinom(48, 100, 0.5)
```

- Graph of the CDF, illustrating how `qbinom()` works.

```{r}
n = 100
p = 0.5

prob3 = tibble(
  x = 0:100,
  prob = dbinom(x, n, p),
  cum_prob = cumsum(prob))

ggplot(prob3, aes(x = x, y = cum_prob)) +
  geom_step() +
  geom_hline(yintercept = 0.4,
             color = "red",
             linetype = "dashed") +
  scale_x_continuous(breaks = 45:55,
                     limits = c(45,55),
                     minor_breaks = NULL)
```

The 0.4 quantile of the $\text{Binomial}(100, 0.5)$ quantile is equal to 49 because:

- $P(X \le 49) \ge 0.4$
- $P(X \ge 49) \ge 1 - 0.4 = 0.6$
- The CDF crosses the horizontal line at $y = 0.4$ at the value $x = 49$.

Here is another graph which shows the graph of the binomial probabilities with the first 0.4 being red and the next 0.6 being blue.

The switch happens at $x = 49$.

```{r}
gbinom(100, 0.5, scale = TRUE) +
  geom_binom_density(100, 0.5, a = 30, b = 48, color = "red") +
  geom_segment(aes(x = 49,
                   xend = 49,
                   y = 0,
                   yend = pbinom(49, 100, 0.5) - 0.4),
               color = "red")
```

### Problems

- $X \sim \text{Binomial}(90, 0.7)$

#### Mean and Variance

```{r}
n = 90
p = 0.7
mu = n*p
mu

sigma = sqrt( n*p*(1-p) )
sigma
```

#### Plot

```{r}
gbinom(n, p, scale = TRUE) +
  geom_vline(xintercept = mu, color = "red") +
  geom_vline(xintercept = mu + c(-1,1)*sigma,
             color = "red",
             linetype = "dashed") +
  geom_vline(xintercept = mu + c(-1,1)*2*sigma,
             color = "red",
             linetype = "dotted")
```

#### Probabilities

- $P(X = \mu)$

```{r}
dbinom(mu, n, p)
```

- $P(\mu - \sigma < X < \mu + \sigma)$

```{r}
mu + sigma
mu - sigma

pbinom(mu + sigma, n, p) - pbinom(mu - sigma, n, p)




pbinom(67, n, p) - pbinom(58, n, p)

pbinom(mu + 2*sigma, n, p) - pbinom(mu - 2*sigma, n, p)
```

#### quantiles

```{r}
qbinom(c(0.05, 0.95), n, p)

gbinom(n, p) +
  geom_binom_density(n, p, b = 56, color = "red") +
  geom_binom_density(n, p, a = 70, color = "red")

pbinom(56,n, p)
```

```{r}
gbinom(n, p) +
  geom_binom_density(n, p, b = qbinom(0.05, n, p),
                     color = "red") +
  geom_binom_density(n, p, a = qbinom(0.95, n, p),
                     color = "red")
```


### Normal

```{r}
ggplot() +
  geom_norm_density(mu = 0, sigma = 1, color = "blue") +
  geom_norm_density(mu = 0, sigma = 0.5, color = "red") +
  geom_norm_density(mu = 2, sigma = 1, color = "darkgreen") +
  labs(
    title = "Three Normal Distributions",
    subtitle = "Blue: N(0, 1). Red: N(0, 0.5). Green: N(2, 1)",
    y = "Density"
  ) +
  geom_hline(yintercept = 0)
```



```{r}
ggplot() +
  geom_norm_fill(mu = 0, sigma = 1,
                 fill = "skyblue") +
  geom_norm_density(0, 1, color = "black") +
  geom_norm_fill(mu = 0, sigma = 1, fill = "blue", alpha = 0.5, a = -0.5, b = 1.5)
```

