---
title: "Section 15: Inference for a Single Mean"
author: "Student's Name Here"
date: "2024-10-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

```{r}
library(tidyverse)
source("../../scripts/ggprob.R")
```

## Sampling distribution of mean

### Data tidying

`TIM.txt` contains data on the times of runners of the Boston Marathon from 2010 and 2011 (and 2013, which we will not analyze).

```{r}
bm_orig <- read_table("../../data/TIM.txt",
                col_types = cols(.default = col_double()))  # Columns are 'double' by default

View(bm_orig)
```

We will make the following changes:

- Eliminate data from 2013
- Create a `Sex` variable with better labels
- Sum the times for racers for different race segments to obtain a total time
- Add an `Age_Range` variable matching the competitor classses in the Boston Marathon.

```{r}
bm <- bm_orig %>% 
  filter(!is.na(`K40-Fin`)) %>% 
  filter(Year < 2013) %>% 
  arrange(Year, BibNum) %>% 
  select(-starts_with("Start"), -HalfMar, -Age2014) %>% 
  
  mutate(Sex = case_when(
    Gender1F2M == 1 ~ "female",
    Gender1F2M == 2 ~ "male")) %>% 
  
  mutate(Time = pmap_dbl(select(., starts_with("K")), sum)) %>% 
  
  mutate(age1 = case_when(
    Age < 35 ~ 18,
    Age >= 80 ~ 80,
    TRUE ~ floor(Age/5)*5),
    age2 = case_when(
      Age < 35 ~ 34,
      TRUE ~ age1 + 4),
    Age_Range = case_when(
      age1 == 80 ~ "80 and older",
      TRUE ~ str_c(age1,"-",age2))) %>% 
  
  select(-Gender1F2M, -age1, -age2) %>% 
  relocate(BibNum, Year, Sex, Age, Age_Range, Time)

bm
```

Each row is a single "run" of the marathon.  If a person competed in 2010 and 2011, they would have two rows.  Columns:

- `BibNum`: bib number, a unique identifier for each competitor within a year
- `Year`: either 2010 or 2011
- `Sex`: female or male
- `Age`: age in years
- `Age_Range`: from 18-34, 35-39, ..., 75-79, and 80 and older
- `Time`: total time to finish in decimal minutes
- `K0-5` through `K40-Fin`: times in decimal minutes for different segments of the race


### Data exploration

Let's look at the age and sex distribution of finishers for each year.

```{r}
bm %>% 
  filter(Year == 2010) %>% 
ggplot(aes(x = Sex, fill = Sex)) +
  geom_bar() +
  xlab("") +
  ylab("# of Finishers") +
  ggtitle("2010 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)
```


```{r}
bm %>% 
  filter(Year == 2011) %>% 
ggplot(aes(x = Sex, fill = Sex)) +
  geom_bar() +
  xlab("") +
  ylab("# of Finishers") +
  ggtitle("2011 Boston Marathon Finishers") +
  facet_wrap(~Age_Range)
```


Let's visualize the sampling distribution of a sample mean.

```{r}
bm_small <- bm %>% 
  filter(Sex == "female", Age_Range == "18-34", Year == 2010)

pretend_pop <- bm_small$Time

ggplot(bm_small) +
  geom_histogram(aes(x = Time), fill = "gray20") +
  labs(title = "Population")
```

```{r}
random_means <- tibble(x = replicate(n = 10000, 
                       mean(sample(pretend_pop, 3557, replace = T)))
                                    )
                       

ggplot(random_means) +
  geom_histogram(aes(x = x), fill = "gray") +
  labs(title = "Sampling Distribution of Sample Mean, n = 3557")
```

It has a normal shape, centered at the original population mean.  In our case, the observed mean is $\bar{x} = 235.5$.

```{r}
bm_summary <- bm %>% 
  filter(Sex == "female", Age_Range == "18-34", Year == 2010) %>% 
  summarize(averageTime = mean(Time), sdTime = sd(Time), n = n())

bm_summary
```


## CI for mean

Let's save the summary values needed to build a CI.

```{r}
xbar <- bm_summary$averageTime
s <- bm_summary$sdTime
n <- bm_summary$n

# Estimated standard error
se <- s/sqrt(n)

# T critical value for 95% confidence
cv <- qt(0.975, df = n - 1)

# 95% CI
c(xbar - cv*se, xbar + cv*se)
```

We can also do this automatically with `t.test`.  The default behavior is to output a 95% CI, but we can change this.

```{r}
t.test(bm_small$Time, conf.level = 0.95)
```


## Hypothesis test for mean

Our p-value is P(T < -7.41) + P(T > 7.41) on the T with n-1 degrees of freedom.

```{r}
test_stat <- (xbar - 240) / (s / sqrt(n))

gt(df = n - 1) +
  geom_vline(xintercept = test_stat, color = "red", lty = 2)
```

Notice the y axis if we zoom in to where the area is visible!

```{r}
gt(df = n - 1, a = -8, b = -7) +
  geom_t_fill(df = n - 1, a = -8, b = test_stat,
              fill = "dodgerblue")
```

Find the p-value with `pt`.

```{r}
2*pt(test_stat, df = n-1)

# More general calculation
left_tail <- pt(-abs(test_stat), df = n-1)
right_tail <- pt(abs(test_stat), df = n-1, lower.tail = FALSE)

left_tail + right_tail
```

We can check our result with `t.test` by specifying the value under the null.

```{r}
t.test(bm_small$Time, conf.level = 0.95)

t.test(bm_small$Time, mu = 240, conf.level = 0.95)
```






