---
title: 'Section 3: Data Types and Structures'
author: "Student's Name Here"
date: "2024-09-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```

## Variables

```{r}
# Initialize with <- or =, but <- is standard R style
x <- 4
my_favorite_number <- 8

# This does not change the value of either x or my_favorite_number
x + my_favorite_number
```

Capitalization matters!

```{r}
x

X
```

```{r}
# This is an illegal variable name
1a <- 3
```


### Numeric objects

```{r}
class(myFavoriteNumber)

class(pi)
```

### Character objects

```{r}
my_name <- "Miranda"

class(my_name)
```

Characters don't obey numeric or logical operators.

```{r}
easy_equation <- "2 + 2"
false_statement <- "5 < 2"

easy_equation
false_statement
```

Some operations do not work with characters.

```{r}
char_year <- "2024"
num_year <- 2024

char_year + 10
num_year + 10
```


### Logical objects

The main objects are `TRUE` and `FALSE`, which we have seen before.

```{r}
# Is three greater than two? Yes, so return TRUE.
3 > 2

# Is two greater than three? No, so return FALSE.
2 > 3
```


## Operators

Mathematical operators let us compute a numeric result with numeric objects.

```{r}
3 + 4
5 - 8

2 * 9.5
6.8 / 2

my_favorite_number ^ 2
my_favorite_number ** 2
```


### Logical operators

Logical operators perform a comparison.

```{r}
my_height <- 60

my_height < 72

my_height < 60
```

```{r}
my_height <= 72
my_height >= 60
```

Use `==` and `!=` to check an exact equality.

```{r}
my_height == 66
my_height != 66
```

The value of `my_height` does not change.

The operators `&` and `|` also perform a comparison.  Instead of comparing numbers, they compare logical statements.

```{r}
# is my_height greater than 72 AND my_height + 6 less than 70?
# one of these fails, so the whole "and" condition fails.
(my_height > 72) & (my_height + 6 < 70)

# is my_height greater than 72 OR my_height + 6 less than 70?
# one of these succeeds, so the whole "or" condition succeeds.
(my_height > 72) | (my_height + 6 < 70)
```


```{r}
TRUE & FALSE
TRUE & TRUE

TRUE | TRUE
TRUE | FALSE
FALSE | FALSE
```


### Missing values

`NA` is the notation for a missing value.

```{r}
nothing <- NA
nothing + 2
```

To check for a missing value, use `is.na`.

```{r}
# Incorrect (R can't interpret this statement):
nothing == NA

# Correct
is.na(nothing)
```

Trying to comopare anything to `NA` returns `NA`.

```{r}
my_height == NA
```


## Functions

```{r}
# To use the sqrt() function, input a numeric value
sqrt(16)

# We need the correct number of arguments
sqrt()
sqrt(9, 16)
```

One of these statements returns an error.  Which one?

```{r}
sqrt(-3)

sqrt("banana")
```

We can also save the results of a function into a variable.

```{r}
# Calculate and store the value
sum_10 <- sum(1, 2, 3, 4, 5, 6, 7, 8, 9, 10) 

# Print it out
sum_10
```

## Vectors and dataframes

A vector is a sequence of the same type of object.  Access an item by index with `[]`.

```{r}
favorite_foods <- c("noodles", "lamb stew", "corn on the cob")
prime_numbers <- c(2, 3, 5, 7, 11)

# Notice how these variables are storing the entire sequence.
favorite_foods
prime_numbers

favorite_foods[2]
prime_numbers[1]
```

If you try to combine different types, R will try to force the objects to be the same type.

```{r}
three_things = c(2, TRUE, "banana")

three_things

class(three_things)
```

### Numeric vector shortcuts

We often want to make a numeric sequence following a specific pattern.  Use `:` to get consecutive numbers and `seq` to get a sequence.

```{r}
1:10
sum(1:10)
```

```{r}
seq(from = 0, to = 10, by = 2)

seq(14, 49, 7)
```

Some functions have default arguments.  The default for `seq` is to start from 1.

```{r}
seq(to = 5, by = 2)
```

Not all arguments have a default value. If you fail to provide an argument which doesn’t have a default, you will get an error.

```{r}
mean(1:10)

some_nums <- c(8, 10, NA)
mean(some_nums)

mean(some_nums, na.rm = TRUE)

# This will give us an error.  Whare are we finding the mean of?
mean(na.rm = TRUE)
```

### Vector operations

Any operation on a vector or multiple vectors will be done individually to the first element of each, then the second element of each, and so on.  We do not need to make a loop to do this type of operation.

```{r}
prime_numbers

prime_numbers + 13

# Use c() to add a new element
c(prime_numbers, 13)
```

This also works for logical operators.

```{r}
prime_numbers == 2

sum(prime_numbers == 2)
```

We can also use `%in%` to look for a specific item.

```{r}
2 %in% prime_numbers
1 %in% prime_numbers
```

We can also do element-wise operations over two vectors.

```{r}
onetwothree <- 1:3
fourfivesix <- 4:6

onetwothree + fourfivesix

fourfivesix - onetwothree

onetwothree * fourfivesix
```

We have to be careful if the vectors are different sizes!

```{r}
fourfive <- 4:5

onetwothree + fourfive
```

### Common vector functions

```{r}
prime_numbers <- c(2, 3, 5, 7, 11)

log(prime_numbers)

min(prime_numbers)
max(prime_numbers)

mean(prime_numbers)
sum(prime_numbers)

median(prime_numbers)
```

Try to experiment with vectors!


### Dataframes

Dataframes are a way to organize multiple vectors.  They're the usual tool for representing real life data.  We can build them with the `tibble` function from tidyverse.

```{r}
# "letters" is a built in R vector with the 26 lowercase letters
letters

library(tidyverse) # Need to do this to get tibble to run

alphabet <- tibble(char = letters,
                   pos = 1:26,
                   is_vowel = letters %in% c("a", "e", "i", "o", "u"))
alphabet
```

Each row represents an individual letter.  We can subset a dataframe by index with the syntax `[,]`.  We can find a row by its name with `$`.

```{r}
# Value of 7th row, 1st column
alphabet[7, 1]

# Ninth row only
alphabet[9, ]

# Third column only
alphabet[ , 3]

alphabet$is_vowel
```

### Real life example

Run the following chunk, which takes data from the Stat 240 course notes.  Make sure you have the tidyverse library loaded.

```{r}
eruptions_recent <- read_csv(
  "https://bwu62.github.io/stat240-revamp/data/eruptions_recent.csv",
  show_col_types = FALSE
)

view(eruptions_recent)
```

Make a new chunk and explore the data with the functions `head`, `glimpse`, `colnames`, `dim`, `nrow`, `ncol`.










